---
title: "Pairwise colocalization with colocPipeline"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pairwise colocalization with colocPipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

`colocPipeline()` runs pairwise QTL ↔ GWAS colocalization using
`coloc::coloc.bf_bf` on the SuSiE log Bayes-factor (LBF) matrices
carried by each fine-mapping entry. For multi-trait colocalization
across many QTL outcomes in one model, use
[`colocboostPipeline`](colocboost-pipeline.html) instead.

Inputs:

- **QTL side** — a `QtlFineMappingResult` produced by
  [`fineMappingPipeline`](fine-mapping.html).
- **GWAS side** — either a `GwasFineMappingResult` (used directly) or a
  `GwasSumStats` (the pipeline fine-maps it inline first).

The output is a long-format data frame with one row per
`(QTL tuple, GWAS tuple, credible-set pair)` combination and the
standard coloc posterior columns `PP.H0.abf` … `PP.H4.abf`.

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled}
data(qtl_dataset_example, gwas_sumstats_s4_example)
qtl_dataset_example      <- fixupExampleGenotypePaths(qtl_dataset_example)
gwas_sumstats_s4_example <- fixupExampleGenotypePaths(gwas_sumstats_s4_example)

# Build a QtlFineMappingResult from the bundled QtlDataset.
qtlFmr <- fineMappingPipeline(qtl_dataset_example, methods = "susie",
                               cisWindow = 1e6)
```

## QTL FMR + GWAS sumstats (inline GWAS fine-mapping)

The simplest call: hand `colocPipeline()` the QTL fine-mapping result
and a GWAS `SumStats`. The pipeline runs `fineMappingPipeline()`
internally on the GWAS sumstats (with `finemappingMethods`), then runs
`coloc.bf_bf` per (QTL, GWAS) pair.

```{r coloc-quickstart}
res <- colocPipeline(
  qtlFineMappingResult = qtlFmr,
  gwasInput            = gwas_sumstats_s4_example,
  finemappingMethods   = "susie")
head(res)
```

Use `returnGwasFineMapping = TRUE` to get the inline GWAS FMR back as an
attribute on the result:

```{r coloc-return-fmr, eval=FALSE}
res2 <- colocPipeline(
  qtlFineMappingResult  = qtlFmr,
  gwasInput             = gwas_sumstats_s4_example,
  returnGwasFineMapping = TRUE)
attr(res2, "gwasFineMapping")
```

## QTL FMR + pre-computed GWAS FMR

If you already have a `GwasFineMappingResult` (e.g. computed once and
reused), pass it directly — the pipeline skips the inline fine-mapping
step:

```{r coloc-precomputed-fmr, eval=FALSE}
gwasFmr <- fineMappingPipeline(gwas_sumstats_s4_example, methods = "susie")
colocPipeline(qtlFineMappingResult = qtlFmr,
              gwasInput            = gwasFmr)
```

## Filtering effects fed into coloc

`coloc::coloc.bf_bf` accepts one LBF matrix per side. By default, every
effect with non-trivial prior variance (`V > priorTol`) gets included.
Two opt-in filters tighten this:

```{r coloc-filters-table, echo=FALSE}
df <- data.frame(
  Knob = c("`filterLbfCs`", "`filterLbfCsSecondary`",
           "`filterLbfCsConcentration`", "`priorTol`"),
  When = c("`filterLbfCsSecondary` is NULL",
           "`filterLbfCsSecondary` is non-NULL (overrides `filterLbfCs`)",
           "only with `filterLbfCsSecondary`",
           "default filter (no `filterLbfCs*`)"),
  What = c("TRUE keeps only effects that produced a credible set (`trimmedFit$sets$cs_index`).",
           "Numeric coverage in (0,1). Runs a CS-concentration filter at that coverage.",
           "Numeric in (0,1). Concentration factor: a CS at coverage X is kept only if it spans fewer than `nVariants * X * concentration` variants.",
           "Drop effects whose estimated prior variance is at or below this threshold."),
  Default = c("FALSE", "NULL", "0.5", "1e-9"),
  stringsAsFactors = FALSE)
knitr::kable(df)
```

```{r coloc-filter-example, eval=FALSE}
colocPipeline(
  qtlFineMappingResult     = qtlFmr,
  gwasInput                = gwas_sumstats_s4_example,
  filterLbfCsSecondary     = 0.5,   # 50% CS concentration filter
  filterLbfCsConcentration = 0.5)   # span < 25% of locus variants
```

## Enrichment-aware coloc (enloc mode)

Pass a per-`(gwasStudy, qtlContext)` enrichment table (the output of
`qtlEnrichmentPipeline()`) to scale the shared-signal prior:

\[
p_{12}^{\text{used}} = \min\bigl(p_{12}\,(1 + \text{enrichment}),\, p_{12,\max}\bigr)
\]

Pairs without a matching enrichment row fall back to the baseline
`p12` with a warning. Two extra columns are added to the output:
`enrichment` (the lookup) and `p12Used` (what was actually passed to
`coloc.bf_bf`).

```{r coloc-enloc, eval=FALSE}
enrichmentTable <- data.frame(
  gwasStudy  = "trait1",
  qtlContext = "brain",
  enrichment = 2.5,
  stringsAsFactors = FALSE)
colocPipeline(
  qtlFineMappingResult = qtlFmr,
  gwasInput            = gwas_sumstats_s4_example,
  enrichment           = enrichmentTable,
  p12Max               = 1e-3)
```

## PIP renormalization on variant intersection

When the QTL and GWAS fine-mapping results were fit on different
variant sets (common when the user declines RAISS imputation, or when
the GWAS FMR carries variants the QTL FMR doesn't), `adjustPips = TRUE`
(default) renormalizes each side's PIPs to the cross-FMR variant
intersection before any per-pair inference. Pass `FALSE` to use the
FMRs as supplied.

```{r coloc-adjustpips, eval=FALSE}
colocPipeline(
  qtlFineMappingResult = qtlFmr,
  gwasInput            = gwas_sumstats_s4_example,
  adjustPips           = TRUE)   # default
```

## LD-sketch consistency

The QTL `FineMappingResult` and the GWAS `SumStats`/FMR must share an
LD sketch (or the QTL side's `ldSketch` must be `NULL` — indicating an
individual-level fit, in which case the check is skipped on the QTL
side). Mismatch is a hard error.

## Common parameters

```{r coloc-params, eval=FALSE}
colocPipeline(
  qtlFineMappingResult     = qtlFmr,
  gwasInput                = gwas_sumstats_s4_example,
  filterLbfCs              = FALSE,
  filterLbfCsSecondary     = NULL,    # e.g. 0.5
  filterLbfCsConcentration = 0.5,
  priorTol                 = 1e-9,
  p1                       = 1e-4,
  p2                       = 1e-4,
  p12                      = 5e-6,
  finemappingMethods       = "susie", # inline GWAS FM when needed
  returnGwasFineMapping    = FALSE,
  enrichment               = NULL,    # data.frame for enloc mode
  p12Max                   = 1e-3,
  adjustPips               = TRUE)
```

## Reading the result

```{r read-result}
str(res, max.level = 1L)
head(res[order(res$PP.H4.abf, decreasing = TRUE), ])
```

Each row is one CS-pair combination; in particular the `idx1` /
`idx2` columns identify the QTL / GWAS effect indices within the
respective LBF matrices, and `PP.H4.abf` is the standard coloc
shared-signal posterior.

## Next steps

- For multi-trait colocalization across many QTL outcomes
  simultaneously, see
  [`colocboostPipeline`](colocboost-pipeline.html).
- Couple coloc evidence with per-gene TWAS Z + MR via
  [`causalInferencePipeline`](twas-zscore.html).
