---
title: "Multi-trait colocalization with ColocBoost"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-trait colocalization with ColocBoost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

`colocboostPipeline()` runs the [colocboost](https://github.com/StatFunGen/colocboost)
multi-trait variable-selection model on QTL data — optionally with one
or more GWAS studies layered on top. It dispatches on the QTL input
class:

| QTL input | What it is |
|---|---|
| `QtlDataset` | Single-study, individual-level: per-context residualized X/Y are extracted from the dataset |
| `QtlSumStats` | QTL summary statistics with a shared LD reference (must be QC'd) |
| `MultiStudyQtlDataset` | One or more individual-level studies + optional embedded `QtlSumStats` |

GWAS input (optional) is always a `GwasSumStats`, passed as the
separate `gwasSumStats` argument. `colocboostPipeline()` does **not**
accept a `FineMappingResult` for either side — colocboost runs its own
variable selection.

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled}
data(qtl_dataset_example,
     qtl_sumstats_example,
     gwas_sumstats_s4_example,
     multi_study_qtl_dataset_example)
qtl_dataset_example             <- fixupExampleGenotypePaths(qtl_dataset_example)
qtl_sumstats_example            <- fixupExampleGenotypePaths(qtl_sumstats_example)
gwas_sumstats_s4_example        <- fixupExampleGenotypePaths(gwas_sumstats_s4_example)
multi_study_qtl_dataset_example <- fixupExampleGenotypePaths(multi_study_qtl_dataset_example)
```

## Three analysis variants

ColocBoost can run three different model flavours in one call.

```{r cb-flags-table, echo=FALSE}
df <- data.frame(
  Flag            = c("`xqtlColoc`",   "`jointGwas`",         "`separateGwas`"),
  Default         = c("TRUE",          "FALSE",                "FALSE"),
  What            = c("QTL-only colocboost across the QTL contexts.",
                      "Single non-focal colocboost model that stacks all QTL contexts/studies with every supplied GWAS study.",
                      "One focal colocboost model per GWAS study (the GWAS is the focal outcome)."),
  stringsAsFactors = FALSE)
knitr::kable(df)
```

## xQTL-only colocalization

The default — run colocboost over the contexts in the QTL data with no
GWAS layer.

```{r xqtl-only}
res <- colocboostPipeline(qtl_dataset_example, xqtlColoc = TRUE)
str(res, max.level = 1L)
```

The result is a named list with four entries:

- `xqtl_coloc` — output of the QTL-only run (NULL when `xqtlColoc = FALSE`)
- `joint_gwas` — output of the joint QTL+GWAS run (NULL when `jointGwas = FALSE`)
- `separate_gwas` — list of per-GWAS focal runs (empty when `separateGwas = FALSE`)
- `computing_time` — per-variant timing record

## QTL + GWAS (joint model)

```{r joint-gwas}
res2 <- colocboostPipeline(
  qtl_dataset_example,
  gwasSumStats = gwas_sumstats_s4_example,
  xqtlColoc    = FALSE,
  jointGwas    = TRUE)
names(res2)
```

## QTL + GWAS (per-GWAS focal models)

When you want one colocboost run per GWAS trait, with that trait as the
focal outcome:

```{r separate-gwas}
res3 <- colocboostPipeline(
  qtl_dataset_example,
  gwasSumStats = gwas_sumstats_s4_example,
  xqtlColoc    = FALSE,
  separateGwas = TRUE)
names(res3$separate_gwas)
```

## Sumstats-only colocalization

A `QtlSumStats` input goes through the same dispatch with no
individual-level data needed; the LD comes from the `ldSketch`
`GenotypeHandle` carried on the collection.

```{r sumstats}
resSs <- colocboostPipeline(
  qtl_sumstats_example,
  gwasSumStats = gwas_sumstats_s4_example,
  xqtlColoc    = FALSE,
  jointGwas    = TRUE)
```

## Multi-study individual-level

```{r multistudy}
resMs <- colocboostPipeline(
  multi_study_qtl_dataset_example,
  gwasSumStats = gwas_sumstats_s4_example,
  xqtlColoc    = TRUE,
  jointGwas    = TRUE)
```

## Focal trait

`focalTrait` makes the xQTL-only run treat one specific trait as the
focal outcome:

```{r focal, eval=FALSE}
colocboostPipeline(
  qtl_dataset_example,
  xqtlColoc  = TRUE,
  focalTrait = "ENSG_example")
```

## Common parameters

```{r cb-params, eval=FALSE}
colocboostPipeline(
  qtlData      = qtl_dataset_example,
  gwasSumStats = gwas_sumstats_s4_example,
  contexts     = "brain",
  traitId      = "ENSG_example",
  region       = NULL,            # alternative to traitId+cisWindow
  cisWindow    = 5e5,             # required with traitId
  focalTrait   = NULL,
  xqtlColoc    = TRUE,
  jointGwas    = FALSE,
  separateGwas = FALSE,
  # Forward to colocboost::colocboost
  M            = 10,
  L            = 5,
  output_level = 1L)
```

Trait / context / region restrictions are pre-construction concerns on
the `QtlDataset` (`keepSamples`, `keepVariants`, the per-context
`SummarizedExperiment` row selection); the pipeline parameters above
are the per-call refinements.

## QC contract

- **Individual-level (`QtlDataset` / `MultiStudyQtlDataset`)**: QC
  (MAF / MAC / X-variance / per-sample missingness, sample / variant
  restrictions) lives on the `QtlDataset` constructor and is applied
  lazily inside `getGenotypes()` / `getResidualizedGenotypes()`. The
  pipeline does not run a separate pass.
- **Summary statistics (`QtlSumStats` / `GwasSumStats`)**: all
  variant filters, panel harmonization, LD-mismatch detection, and
  RAISS imputation live in [`summaryStatsQc()`](rss-qc.html). The
  pipeline rejects inputs whose `getQcInfo()` is empty.

## Next steps

- For pairwise QTL ↔ GWAS coloc with the classic
  `coloc.bf_bf` model (instead of multi-trait ColocBoost), see
  [`colocPipeline`](coloc-pipeline.html).
- To compose colocboost results with a per-gene TWAS Z + MR analysis,
  see [`causalInferencePipeline`](twas-zscore.html).
