---
title: "Fine-mapping QTLs and GWAS with the pecotmr SuSiE Wrapper"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fine-mapping QTLs and GWAS with the pecotmr SuSiE Wrapper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

Fine-mapping reduces a region with one or more associations to a small set
of credible causal variants. `pecotmr` wraps
[SuSiE](https://stephenslab.github.io/susieR/) (Sum of Single Effects) and
SuSiE-RSS (its summary-statistics analog) and adds:

- A **two-stage fit** — SuSiE-inf first, then SuSiE initialized from the
  SuSiE-inf result — that improves credible-set purity in regions with
  background polygenic signal.
- Unified **post-processing** producing the same 22-column `top_loci`
  table regardless of whether the input was individual-level data or
  summary statistics, and regardless of whether you fit SuSiE, SuSiE-inf,
  SuSiE-RSS, fSuSiE, or mvSuSiE.
- **Pipeline-level entry points** that bundle loading, QC, fine-mapping,
  and (for individual-level data) TWAS-weight learning.

This vignette demonstrates all three workflows:

| Data you have | Entry point | Example dataset |
|---|---|---|
| Individual-level genotype + phenotype (QTL studies) | `univariate_analysis_pipeline()` | `eqtl_region_example` |
| QTL summary statistics + LD reference | `rss_analysis_pipeline()` | placeholder (real QTL nominal output) |
| GWAS summary statistics + LD reference | `rss_analysis_pipeline()` | placeholder (real GWAS sumstats) |

For SuSiE theory, see the
[susieR vignettes](https://stephenslab.github.io/susieR/articles/index.html);
this vignette focuses on how to drive it through `pecotmr`.

```{r load-packages}
library(pecotmr)
```

## Quick start: individual-level fine-mapping

`eqtl_region_example` contains a genotype matrix `X` (415 samples ×
2,828 variants) and a residualized molecular phenotype `y_res`. Below
we run the full pipeline — initial PIP-screening, SuSiE-inf, SuSiE
initialised from SuSiE-inf, post-processing, and (optionally)
TWAS-weight learning — in a single call.

```{r individual-quickstart}
data(eqtl_region_example)
X <- eqtl_region_example$X
y <- as.numeric(eqtl_region_example$y_res)
maf <- apply(X, 2, function(g) {
  f <- mean(g, na.rm = TRUE) / 2
  min(f, 1 - f)
})

# Subset to a smaller window for a fast vignette build; the same call works
# on the full region.
set.seed(42)
sub <- sort(sample(ncol(X), 500))

fit <- univariate_analysis_pipeline(
  X = X[, sub], Y = y, maf = maf[sub],
  region = "chr22:32119000-33000000",
  twas_weights = FALSE,
  finemapping_extra_opts = list(refine = FALSE),
  verbose = 0
)
```

The return value is a list with the fitted models and the unified
`top_loci` table:

```{r individual-results}
names(fit)
dim(fit$top_loci)
head(fit$top_loci[, c("variant", "pip", "cs_95", "cs_70",
                       "cs_95_purity", "method")], 5)
```

The methods column shows which model each row came from — the pipeline
runs both SuSiE-inf and SuSiE by default and row-binds their top loci.
For most downstream consumers, the SuSiE rows are what you want; SuSiE-inf
rows are useful as a sanity check on inferred non-sparse effects.

## Choosing among SuSiE variants

`pecotmr` exposes five SuSiE-family fine-mapping variants. Pick which
to fit via the `methods` parameter on `univariate_analysis_pipeline()`
(individual-level) or `rss_analysis_pipeline()` / `susie_rss_pipeline()`
(summary statistics):

| Variant | Individual-level | Summary statistics | What it models |
|---|---|---|---|
| **SuSiE alone** | `methods = "susie"` | `methods = "susie_rss"` | Sparse causal effects only. No infinitesimal/background term. Lightest weight; can fold polygenic background into noisy "extra" credible sets. |
| **SuSiE with SuSiE-inf init** (default) | `methods = c("susie_inf", "susie")` + `add_susie_inf = TRUE` | `methods = c("susie_inf_rss", "susie_rss")` + `add_susie_inf = TRUE` | Two-stage: SuSiE-inf absorbs background polygenicity first, SuSiE then refines sparse signals from that initialisation. Recommended default for most QTL and GWAS fine-mapping. |
| **SuSiE-inf alone** | `methods = "susie_inf"` | `methods = "susie_inf_rss"` | Sparse + infinitesimal joint fit. Useful for regions with strong polygenic background where the SuSiE refinement step is undesired. |
| **SuSiE-ash alone** | `methods = "susie_ash"` | `methods = "susie_ash_rss"` | Sparse + adaptive shrinkage (ASH) mixture. Useful when you suspect a mix of medium-effect and small-effect variants beyond a small handful of large ones. |
| **SuSiE-ash with SuSiE-inf init** | `methods = c("susie_inf", "susie_ash")` + `add_susie_inf = TRUE` | `methods = c("susie_inf_rss", "susie_ash_rss")` + `add_susie_inf = TRUE` | Same chained two-stage idea as the SuSiE variant but with ASH on the downstream fit: SuSiE-inf absorbs polygenic background first, then SuSiE-ash refines using the mixture prior. |

The chained SuSiE-inf → SuSiE two-stage is the default both for
backward compatibility and because it produces clean credible sets in
most settings: SuSiE alone can fold polygenic background into noisy
"extra" credible sets; SuSiE-inf absorbs that background first so the
SuSiE refinement concentrates on sparse signals.

`add_susie_inf = TRUE` chains SuSiE-inf into **any** of `"susie"` or
`"susie_ash"` that appear in `methods`. Listing all three
(`methods = c("susie_inf", "susie", "susie_ash")` with
`add_susie_inf = TRUE`) fits SuSiE-inf once, then both SuSiE and
SuSiE-ash initialised from it — useful for direct head-to-head
comparison on the same warm start.

All five populate the same 22-column `top_loci` table. When you fit
multiple methods in one call, the rows are row-bound and labelled by
the `method` column.

> Note: SuSiE-inf chain init only carries information when the
> SuSiE-inf fit finds non-zero mappable effects. On a region where
> SuSiE-inf explains everything via its infinitesimal component (all
> `V == 0`, all `mu == 0`), upstream `susieR` correctly treats the
> init as empty, and the chained downstream fit collapses to the
> independent run. This is the right behaviour, not a bug — but it
> means head-to-head differences only show up on regions with at
> least one sparse signal.

### Backward compatibility with `add_susie_inf`

The pre-existing `add_susie_inf` boolean still works exactly as before
when `methods` is `NULL` (the default):

- `add_susie_inf = TRUE` (default) → fit SuSiE-inf, then chain it into
  SuSiE. Equivalent to `methods = c("susie_inf", "susie")`.
- `add_susie_inf = FALSE` → plain SuSiE alone. Equivalent to
  `methods = "susie"`.

When `methods` is set explicitly, `add_susie_inf` controls whether the
chained-init shortcut is applied to any `"susie"` or `"susie_ash"`
method that appears alongside `"susie_inf"`. Set `add_susie_inf = FALSE`
to fit each requested method independently rather than chained.

### Fitting multiple variants in one call

```{r multi-method-fit, eval=FALSE}
multi <- univariate_analysis_pipeline(
  X = X[, sub], Y = y, maf = maf[sub],
  methods = c("susie_inf", "susie", "susie_ash"),
  twas_weights = FALSE,
  signal_cutoff = 0
)
# top_loci has rows for all three methods; filter on `method` to inspect each
table(multi$top_loci$method)
```

### Lower-level entry points

If you want the raw fits without `pecotmr`'s post-processing:

```{r two-stage-fit, eval=FALSE}
# Two-stage convenience for the recommended default:
fits <- fit_susie_inf_then_susie(X[, sub], y)         # list(susie, susie_inf)
fits_rss <- fit_susie_inf_then_susie_rss(z, R, n)

# Or call susieR directly with the right unmappable_effects flag:
inf_fit <- susieR::susie(X, y, unmappable_effects = "inf",
                          convergence_method = "pip", refine = FALSE)
ash_fit <- susieR::susie(X, y, unmappable_effects = "ash",
                          convergence_method = "pip")

# Then post-process any combination through the unified pipeline:
post <- postprocess_finemapping_fits(
  fits = list(susie_inf = inf_fit, susie_ash = ash_fit),
  data_x = X[, sub], data_y = y, maf = maf[sub],
  coverage = 0.95, secondary_coverage = c(0.7, 0.5)
)
out <- format_finemapping_output(post, primary_method = "susie_inf")
```

## The unified `top_loci` table

`postprocess_finemapping_fits()` produces a single table with the
following 22 columns, in this order:

| #  | Column                  | Source                                                 |
|---:|-------------------------|--------------------------------------------------------|
|  1 | `#chr`                  | parsed from `variant`                                  |
|  2 | `start`                 | `pos - 1` (BED 0-based)                                |
|  3 | `end`                   | `pos` (BED half-open)                                  |
|  4 | `a1`                    | from parsed `variant`                                  |
|  5 | `a2`                    | from parsed `variant`                                  |
|  6 | `variant`               | the row's variant id (`chr:pos:A2:A1`)                 |
|  7 | `gene`                  | phenotype identifier (`colnames(data_y)[1]`)           |
|  8 | `event`                 | `paste(condition_id, gene, sep = "_")`                 |
|  9 | `n`                     | analyzed sample count                                  |
| 10 | `maf`                   | from the `maf` argument                                |
| 11 | `beta`                  | univariate effect (`sumstats$betahat`)                 |
| 12 | `se`                    | univariate SE (`sumstats$sebetahat`)                   |
| 13 | `pip`                   | per-variant SuSiE PIP                                  |
| 14 | `posterior_effect_mean` | `colSums(alpha * mu)`                                  |
| 15 | `posterior_effect_se`   | `sqrt(pmax(colSums(alpha * mu2) - mean^2, 0))`         |
| 16 | `cs_95`                 | CS membership at coverage 0.95 (see below)             |
| 17 | `cs_70`                 | CS membership at coverage 0.70                         |
| 18 | `cs_50`                 | CS membership at coverage 0.50                         |
| 19 | `cs_95_purity`          | 0.95-coverage CS purity for this row                   |
| 20 | `method`                | `"susie"`, `"susie_inf"`, `"susie_rss"`, ...           |
| 21 | `grange_start`          | parsed from the `region` argument                      |
| 22 | `grange_end`            | parsed from the `region` argument                      |

### CS membership strings

`cs_95`, `cs_70`, `cs_50` are **character strings**, not integers. Each
value is `<method>_<cs_index>` where `<cs_index>` is the per-method
1-based credible-set number:

- `susie_1`, `susie_2`, ... for SuSiE credible sets
- `susie_inf_1`, ... for SuSiE-inf credible sets
- `<method>_0` is the sentinel for "PIP-only retention, not in any CS at
  this coverage"

CS indices are **not** sequenced across methods — each method numbers
its credible sets from 1.

### Row uniqueness

The row identity is `(variant, gene, method, cs_membership)`:

- A variant in CS 1 and CS 2 of the same method (overlapping CS) gives
  **two** rows.
- A variant in both `susie` CS 1 and `susie_inf` CS 1 gives **two** rows
  with the same `(variant, gene)` but different `method`.
- A PIP-retained variant not in any CS gives **one** row per method with
  `cs_* = "<method>_0"` and `cs_95_purity = 0`.

This shape preserves overlapping-CS evidence both within and across
methods. Downstream consumers (xqtl-protocol, transMap, ColocBoost,
TWAS) read the columns above; the schema is the contract.

## QTL summary-statistics fine-mapping

When you have z-scores instead of `X` and `Y`, use
`rss_analysis_pipeline()`. It bundles:

1. `load_rss_data()` — tabix-region-extracts the sumstats and
   standardises column names via a YAML mapping (see the
   [RSS QC vignette](rss-qc.html) for the YAML schema).
2. `summary_stats_qc()` — runs allele harmonization, SLALOM (default)
   or DENTIST QC, and optional RAISS imputation. The vignette referenced
   above covers QC in detail.
3. `susie_rss_pipeline()` — fits the two-stage SuSiE-RSS model and
   post-processes into the unified `top_loci` table.

### Pipeline call (file-based, production shape)

```{r qtl-rss-pipeline, eval=FALSE}
# TODO(data): replace with real QTL sumstats and LD-metadata paths
qtl_sumstat_path  <- "<path/to/qtl_nominal.tsv.gz>"
qtl_column_map    <- "<path/to/qtl_column_map.yml>"
ld_meta_path      <- "<path/to/ld_meta.tsv>"
region            <- "<chr:start-end>"
gene_id           <- "<ENSG...>"
n_qtl_sample      <- 0  # 0 = read from sumstats

ld_data <- load_LD_matrix(ld_meta_path, region = region)

fm <- rss_analysis_pipeline(
  sumstat_path        = qtl_sumstat_path,
  column_file_path    = qtl_column_map,
  LD_data             = ld_data,
  n_sample            = n_qtl_sample,
  region              = region,
  extract_region_name = gene_id,
  region_name_col     = 4L,
  qc_method           = "slalom",
  finemapping_method  = "susie_rss",
  finemapping_opts    = list(L = 20, L_greedy = 5,
                             coverage = c(0.95, 0.7, 0.5),
                             signal_cutoff = 0.025,
                             min_abs_corr = 0.8),
  impute              = TRUE
)
head(fm$rss_data_analyzed[[1]]$top_loci)
```

### Lower-level alternative (in-memory, runnable here)

For demonstration on the shipped example data, we skip the file-based
loading and use `susie_rss_pipeline()` directly with the sumstats and
LD already in memory. This is what `rss_analysis_pipeline()` calls
internally after QC.

```{r qtl-rss-inmemory}
data(gwas_sumstats_example)
data(eqtl_region_example)

# We treat the shipped GWAS-style sumstats as if they were QTL z-scores
# for this demonstration. In a real workflow you would load QTL nominal
# output via load_rss_data().
sumstats <- gwas_sumstats_example
X_ref    <- eqtl_region_example$X
R        <- compute_LD(X_ref, method = "sample")
R_sub    <- R[sumstats$variant_id, sumstats$variant_id]

fm_rss <- susie_rss_pipeline(
  sumstats          = sumstats,
  LD_mat            = R_sub,
  n                 = nrow(X_ref),
  L                 = 10, L_greedy = 5,
  analysis_method   = "susie_rss",
  coverage          = 0.95,
  secondary_coverage = c(0.7, 0.5)
)
names(fm_rss)
dim(fm_rss$top_loci)
head(fm_rss$top_loci[, c("variant", "pip", "cs_95",
                          "cs_95_purity", "method")], 5)
```

Note that the output schema is the same 22 columns as the
individual-level pipeline — `method` is `"susie_rss"` instead of
`"susie"`, but everything downstream (ColocBoost, xqtl-enrichment,
TWAS) consumes the same table.

## GWAS summary-statistics fine-mapping

GWAS fine-mapping uses the **same** `rss_analysis_pipeline()` /
`susie_rss_pipeline()` call. The only differences are at the loading
layer:

- **Column mapping.** GWAS conventions (effect/non-effect, OR vs beta,
  per-variant `n`) differ from QTL nominal output. The
  [RSS QC vignette](rss-qc.html) shows the YAML schema; common GWAS
  alternatives are `effect_allele`/`other_allele` for `A1`/`A2`,
  `OR` (you'll need `log(OR)` for `beta`), and `N` for `n_sample`.
- **No `extract_region_name`.** GWAS sumstats are not gene-keyed, so
  you slice by physical region only. Pass `region = "chr:start-end"`
  and omit `extract_region_name` / `region_name_col`.
- **Per-variant sample size.** GWAS sumstats often have a per-variant
  `n_sample` column (some variants were imputed in fewer cohorts);
  `load_rss_data()` reads that automatically when the column is mapped.

```{r gwas-rss-pipeline, eval=FALSE}
# TODO(data): replace with real GWAS sumstats and LD-metadata paths
gwas_sumstat_path <- "<path/to/gwas.tsv.gz>"
gwas_column_map   <- "<path/to/gwas_column_map.yml>"
ld_meta_path      <- "<path/to/ld_meta.tsv>"
region            <- "<chr:start-end>"

ld_data <- load_LD_matrix(ld_meta_path, region = region)

fm_gwas <- rss_analysis_pipeline(
  sumstat_path       = gwas_sumstat_path,
  column_file_path   = gwas_column_map,
  LD_data            = ld_data,
  n_case             = 0,        # set if known; 0 reads from sumstats
  n_control          = 0,
  region             = region,
  qc_method          = "slalom",
  finemapping_method = "susie_rss",
  impute             = TRUE
)
```

The shipped `gwas_finemapping_example` dataset shows what the output
looks like for a real GWAS region:

```{r gwas-example-shape}
data(gwas_finemapping_example)
str(gwas_finemapping_example, max.level = 2)
length(gwas_finemapping_example[[1]]$pip)
sum(gwas_finemapping_example[[1]]$pip > 0.1)
```

`gwas_finemapping_example` ships in the legacy `susie_result_trimmed`
shape used by older pecotmr pipelines (PIPs, alpha matrix, credible
sets). New workflows should consume the 22-column `top_loci` table
produced by `rss_analysis_pipeline()` instead.

<!-- TODO(data): once a real GWAS sumstat + LD example is ready, replace
the above with a runnable end-to-end demonstration. -->

## Downstream consumers

The `top_loci` table feeds directly into several other `pecotmr`
workflows:

- **Enrichment** (`xqtl_enrichment_wrapper`): computes prior enrichment
  of QTL fine-mapping signals at GWAS PIPs. See the
  [enrichment vignette](xqtl-enrichment.html).
- **Colocalization** (`colocboost_pipeline`): uses the trimmed SuSiE
  fits — `finemapping_result` slot — as colocalization input. See the
  [ColocBoost vignette](colocboost-pipeline.html).
- **TWAS** (`twas_weights_pipeline`, `twas_pipeline`): uses the SuSiE
  effect estimates as weight learners for transcriptome-wide
  association testing. See the
  [TWAS weights vignette](twas-weights.html) and the
  [TWAS Z-score vignette](twas-zscore.html).

## Summary

| Function | Use it when |
|---|---|
| `univariate_analysis_pipeline()` | Individual-level data: you have `X` (samples × variants), `Y` (phenotype). Pick variant(s) via `methods = c("susie", "susie_inf", "susie_ash")`. |
| `rss_analysis_pipeline()` | Summary statistics (QTL or GWAS): you have z-scores in a tabix file and an LD reference. Pick variant(s) via `methods = c("susie_rss", "susie_inf_rss", "susie_ash_rss")`. |
| `susie_rss_pipeline()` | Lower-level: you already have in-memory `z`, `R` (or `X`), `n`. Same `methods` argument as `rss_analysis_pipeline()`. |
| `fit_susie_inf_then_susie()` / `fit_susie_inf_then_susie_rss()` | Lowest-level convenience for the two-stage chained default (SuSiE-inf init → SuSiE). |
| `postprocess_finemapping_fits()` + `format_finemapping_output()` | Post-process any combination of `susie`, `susie_inf`, `susie_ash`, `susie_rss`, `susie_inf_rss`, `susie_ash_rss`, `mvsusie`, `fsusie` fits into the unified 22-column `top_loci`. |

All five entry points produce — or can be post-processed into — the
same 22-column `top_loci` table documented above, which is the contract
all downstream `pecotmr` consumers read.

```{r session-info}
sessionInfo()
```
