---
title: "Fine-mapping with the pecotmr S4 pipeline"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fine-mapping with the pecotmr S4 pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

`fineMappingPipeline()` is a single S4 generic that dispatches on the
input class:

| Input class | What it represents | Methods supported |
|---|---|---|
| `QtlDataset` | Single-study individual-level data: genotypes + per-context phenotypes | susie, susieInf, susieAsh, mvsusie, fsusie |
| `MultiStudyQtlDataset` | Multiple `QtlDataset`s (optionally with an embedded `QtlSumStats`) | Same as above; runs per study |
| `QtlSumStats` | QTL summary statistics keyed by `(study, context, trait)` | susie, susieInf, susieAsh (RSS implementation auto-dispatched) |
| `GwasSumStats` | GWAS summary statistics keyed by `study` | susie, susieInf, susieAsh (RSS implementation auto-dispatched) |

Method tokens are the same across input classes; `fineMappingPipeline()`
routes `methods = "susie"` to `susieR::susie` for individual-level data
and to `susieR::susie_rss` for summary-statistics inputs automatically.

The output is always a `FineMappingResult` (a DFrame-subclass collection
of per-tuple `FineMappingEntry` payloads).

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled-inputs}
data(qtl_dataset_example,
     qtl_sumstats_example,
     gwas_sumstats_s4_example,
     multi_study_qtl_dataset_example)
# The bundled objects carry relative GenotypeHandle paths; this helper
# resolves them to the installed inst/extdata location.
qtl_dataset_example             <- fixupExampleGenotypePaths(qtl_dataset_example)
qtl_sumstats_example            <- fixupExampleGenotypePaths(qtl_sumstats_example)
gwas_sumstats_s4_example        <- fixupExampleGenotypePaths(gwas_sumstats_s4_example)
multi_study_qtl_dataset_example <- fixupExampleGenotypePaths(multi_study_qtl_dataset_example)

qtl_dataset_example
```

The `*_sumstats_example` objects ship with a non-empty `qcInfo` slot so
they bypass the `summaryStatsQc()` gate. In a real analysis you'd
construct the raw `QtlSumStats` / `GwasSumStats` from your data, then
run `summaryStatsQc()` to populate that slot (see the
[summary-statistics QC vignette](rss-qc.html)).

## Individual-level fine-mapping

`fineMappingPipeline()` always needs to know which genotype window to
fit on. With an individual-level `QtlDataset`, pass `cisWindow` (in
basepairs around each trait's TSS), supply an explicit `region`, or
set both per-call.

```{r individual-fm}
fmr <- fineMappingPipeline(qtl_dataset_example,
                            methods   = "susie",
                            cisWindow = 1e6)
fmr
```

Per-tuple PIPs:

```{r individual-pips}
pip <- getPip(fmr, study = "study1", context = "brain",
              trait = "ENSG_example", method = "susie")
head(sort(pip, decreasing = TRUE), 5)
```

Credible sets:

```{r individual-cs}
getCs(fmr, study = "study1", context = "brain",
      trait = "ENSG_example", method = "susie", coverage = 0.95)
```

The `top_loci` table consolidates per-method PIPs, CS membership, and
sumstats into one long-format data.frame:

```{r individual-toploci}
head(getTopLoci(fmr, study = "study1", context = "brain",
                trait = "ENSG_example", method = "susie"), 5)
```

## Multi-study fine-mapping

A `MultiStudyQtlDataset` runs the per-study pipeline in turn and
concatenates the results into a single `QtlFineMappingResult`:

```{r multistudy-fm}
msFmr <- fineMappingPipeline(multi_study_qtl_dataset_example,
                              methods   = "susie",
                              cisWindow = 1e6)
table(msFmr$study)
```

## QTL summary-statistics fine-mapping

```{r sumstats-qtl}
fmrSs <- fineMappingPipeline(qtl_sumstats_example, methods = "susie")
fmrSs
```

The `QtlSumStats` collection carries an `ldSketch` `GenotypeHandle`; the
pipeline pulls the per-block LD from there automatically.

## GWAS summary-statistics fine-mapping

```{r sumstats-gwas}
fmrGwas <- fineMappingPipeline(gwas_sumstats_s4_example,
                                methods = "susie")
fmrGwas
```

The GWAS path is essentially the QTL-sumstats path with the 1-tuple
`(study)` key instead of the 4-tuple `(study, context, trait, method)`.

## Common parameters

```{r fm-params, eval=FALSE}
fineMappingPipeline(
  qtl_dataset_example,
  methods            = c("susie", "susieInf"),
  contexts           = "brain",
  traitId            = "ENSG_example",
  region             = "chr22:25000000-26000000",
  cisWindow          = 5e5,
  coverage           = 0.95,
  secondaryCoverage  = c(0.7, 0.5),
  signalCutoff       = 0.025,
  minAbsCorr         = 0.8,
  addSusieInf        = TRUE,
  naAction           = "drop",   # "drop" (default) or "impute"
  verbose            = 1)
```

- `naAction` controls how phenotype NAs are handled by
  `getResidualizedPhenotypes()`: `"drop"` removes samples with any NA
  across the requested traits, `"impute"` mean-imputes each trait
  independently.
- `addSusieInf = TRUE` runs SuSiE-inf first and uses its result to
  initialise the susie fit, which improves credible-set purity in
  regions with background polygenic signal.
- `secondaryCoverage` adds secondary CS columns at the requested
  coverages (alongside the primary `coverage`); useful when you want
  both narrow and broad CS reports.

## Joint multi-context / multi-trait fits

`jointSpecification` triggers the dispatch for cross-axis joint
methods (`mvsusie` for cross-context, `fsusie` for ordered-trait):

```{r fm-joint, eval=FALSE}
fineMappingPipeline(
  multi_study_qtl_dataset_example,
  methods            = "mvsusie",
  jointSpecification = list(axis = "context",
                             contexts = c("brain", "liver")))
```

See `?jointSpecification` for the spec grammar.

## Reading the result

`FineMappingResult` is a DFrame subclass: subsetting with `[`,
concatenating with `c()`, and column access via `$` all work as on a
DataFrame. The per-tuple payload lives in the `entry` column as a
`FineMappingEntry`:

```{r fm-entry}
entry <- fmr$entry[[1L]]
class(entry)
slotNames(entry)
```

Each entry carries:

- `variantIds` — the LD-aligned variant set the fit was computed on
- `trimmedFit` — a slim list containing `alpha`, `lbf_variable`, `mu`,
  `mu2`, `X_column_scale_factors`, `pip`, and credible-set metadata
- `topLoci` — the long-format per-variant summary used by downstream
  pipelines and `vcfWriter`
- `sumstats` — optional sumstats passed through for reporting

## Next steps

- Hand a `QtlFineMappingResult` to
  [`colocPipeline`](coloc-pipeline.html) for cross-study colocalization.
- Pass `fineMappingPipeline()` output to
  [`twasWeightsPipeline`](twas-weights.html) so the susie weights are
  contributed to the TWAS ensemble alongside other weight methods.
- Run the full
  [`causalInferencePipeline`](twas-zscore.html) (TWAS + MR) on top of a
  fine-mapping result and GWAS sumstats.
