---
title: "Quality Control for GWAS and QTL Summary Statistics"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Control for GWAS and QTL Summary Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

Regression-with-summary-statistics (RSS) methods — SuSiE-RSS, ColocBoost,
DENTIST, lassosum, PRS-CS, SDPR — all require a summary-statistics table
(`z` or `beta`/`se`) and an LD reference panel that agree about which
variant is which. Two things go wrong in practice:

1. **Allele alignment.** Sumstats and LD panel may flip A1/A2, encode
   strand-ambiguous variants inconsistently, or contain indels and
   duplicates that need filtering.
2. **LD-summary-statistic mismatch.** Even when alleles align, individual
   z-scores can be inconsistent with the LD pattern because of imputation
   errors, ancestry differences between the GWAS and LD panel, or
   per-variant sample-size differences.

`pecotmr` exposes both stages through pipeline-level wrappers:

| Stage | Function | What it does |
|---|---|---|
| 1 — Allele alignment | `rssBasicQc()` | Match alleles, flip effect direction, drop indels/duplicates/strand-ambiguous variants, optionally exclude skip regions. |
| 2 — Mismatch detection | `ldMismatchQc()` / `summaryStatsQc()` | Flag LD-inconsistent variants via SLALOM (default) or DENTIST; optionally re-impute with RAISS. |

The same wrappers work for both GWAS and QTL summary statistics — only the
loading layer differs (QTL sumstats are typically tabix-indexed per-gene
tables that need a column-mapping YAML). The fine-mapping and
TWAS-weights pipelines (`rssAnalysisPipeline()`,
`twasWeightsSumstatPipeline()`) call these QC functions internally with
`qcMethod = "slalom"` as the default; this vignette shows what they do
under the hood so you can run QC standalone or tune its behavior.

```{r load-packages}
library(pecotmr)
```

## Quick start on shipped example data

The `gwas_sumstats_example` dataset is a de-identified GWAS summary table
on a single region, and `eqtl_region_example$X` is the matching
individual-level genotype matrix that we can use as an in-sample LD
reference. The two datasets share all 2,828 variants, so we can focus on
QC mechanics without dealing with cross-cohort variant matching yet.

```{r load-toy}
data(gwas_sumstats_example)
data(eqtl_region_example)

sumstats <- gwas_sumstats_example
X_ref <- eqtl_region_example$X

cat(sprintf("Sumstats: %d variants\nReference panel: %d samples x %d variants\n",
            nrow(sumstats), nrow(X_ref), ncol(X_ref)))
head(sumstats, 3)
```

The shipped sumstats are already aligned to the LD panel (same `variant_id`
strings, same allele orientation). That lets us go straight to the
LD-mismatch step; we will come back to allele harmonization in the next
section.

```{r quick-qc}
R <- computeLd(X_ref, method = "sample")

qc <- ldMismatchQc(
  zScore = sumstats$z,
  R = R[sumstats$variant_id, sumstats$variant_id],
  method = "slalom"
)

cat(sprintf("SLALOM flagged %d / %d variants (%.1f%%)\n",
            sum(qc$outlier, na.rm = TRUE),
            nrow(qc),
            100 * mean(qc$outlier, na.rm = TRUE)))
```

`ldMismatchQc()` is the lowest-level entry point: pass z-scores and an
LD matrix, get back a per-variant `outlier` flag. The pipeline functions
in `pecotmr` call this internally on each region after allele alignment.

## Pipeline integration: `rssBasicQc()` + `summaryStatsQc()`

In a typical workflow you would load LD from a metadata TSV
(`loadLdMatrix()` returns an `LdData` S4 object) and pass it through the
two-stage QC pipeline. We construct an `LdData` from the shipped genotype
matrix to keep the example self-contained.

```{r build-lddata}
ref_panel <- parseVariantId(colnames(X_ref))
ref_panel$variant_id <- colnames(X_ref)

variants_gr <- GenomicRanges::GRanges(
  seqnames = ref_panel$chrom,
  ranges = IRanges::IRanges(start = ref_panel$pos, width = 1L)
)
S4Vectors::mcols(variants_gr) <- S4Vectors::DataFrame(
  variant_id = ref_panel$variant_id,
  A1 = ref_panel$A1,
  A2 = ref_panel$A2
)
block_meta <- data.frame(
  block_id = 1L,
  chrom = ref_panel$chrom[1],
  block_start = min(ref_panel$pos),
  block_end = max(ref_panel$pos),
  size = nrow(ref_panel),
  start_idx = 1L,
  end_idx = nrow(ref_panel)
)

ld_data <- LdData(
  correlation = R,
  variants = variants_gr,
  blockMetadata = block_meta,
  nRef = nrow(X_ref)
)
```

### Stage 1 — `rssBasicQc()`

`rssBasicQc()` performs allele-level harmonization between the sumstats
and the LD panel. It flips signs on `z` (and `beta` if present) for
variants where A1/A2 swap, drops strand-ambiguous A/T and C/G variants,
removes duplicates, and optionally removes indels.

```{r rss-basic-qc}
qc_basic <- rssBasicQc(sumstats, LD_data = ld_data)
sumstats_aligned <- getrssinput(qc_basic)$sumstats
cat(sprintf("After allele QC: %d variants\n", nrow(sumstats_aligned)))
```

The return value is a `QcResult` S4 object; use `getrssinput()$sumstats`
for the cleaned sumstats and `getlddata()` for the LD subset aligned to
those variants.

### Stage 2 — `summaryStatsQc()`

`summaryStatsQc()` dispatches to SLALOM (default) or DENTIST to flag
LD-inconsistent variants, then subsets sumstats and LD to the
non-outliers. With `impute = TRUE` it also runs RAISS to re-impute the
outlier z-scores from the surviving variants.

```{r summary-stats-qc}
qc_full <- summaryStatsQc(
  sumstats = sumstats_aligned,
  LD_data = getlddata(qc_basic),
  method = "slalom"
)

sumstats_clean <- getrssinput(qc_full)$sumstats
cat(sprintf("After LD-mismatch QC: %d variants (%d flagged)\n",
            nrow(sumstats_clean),
            getoutliernumber(qc_full)))
```

`sumstats_clean` is now ready to feed into any downstream RSS method:
`susie_rss()`, `colocboostPipeline()`, `otters_weights()`, etc. The
fine-mapping and TWAS-weights vignettes pick up from here.

## QC for QTL summary statistics

QTL sumstats (e.g. eQTL, sQTL nominal output from tensorQTL or APEX) need
the same two-stage QC. The only practical differences are:

- **Loading.** QTL sumstats are typically very large per-tissue tables,
  so you load them region-by-region via tabix using `loadRssData()`.
- **Column mapping.** Nominal-output column names vary by tool. A small
  YAML file maps your column names to the standardized names
  `pecotmr` expects.

The same `rssBasicQc()` / `summaryStatsQc()` calls work on the
loaded sumstats; the pipeline functions `rssAnalysisPipeline()` and
`twasWeightsSumstatPipeline()` wrap all of this.

### Column mapping

The mapping YAML maps your column names (right) to the standard names
`pecotmr` expects (left). Required keys are `chrom`, `pos`, `A1`, `A2`,
plus either `z` or `beta`/`sebeta`.

```yaml
# TODO(data): replace with the column names from your QTL tool's output
chrom: chromosome
pos: position
A1: alt_allele
A2: ref_allele
beta: slope
sebeta: slope_se
n_sample: ma_count   # optional; per-variant sample size column
```

### Loading and QC

```{r qtl-qc, eval=FALSE}
# TODO(data): replace with real QTL sumstats and LD-metadata paths
qtl_sumstat_path  <- "<path/to/qtl_nominal.tsv.gz>"
qtl_column_map    <- "<path/to/qtl_column_map.yml>"
ld_meta_path      <- "<path/to/ld_meta.tsv>"
region            <- "<chr:start-end>"
gene_id           <- "<ENSG...>"
n_qtl_sample      <- 0  # 0 = read from sumstats; otherwise pass explicit n

# Load sumstats for one gene region, with column standardization
rss_input <- loadRssData(
  sumstat_path        = qtl_sumstat_path,
  column_file_path    = qtl_column_map,
  region              = region,
  extract_region_name = gene_id,
  region_name_col     = 4L,
  n_sample            = n_qtl_sample
)

# Load LD reference for the same region
ld_data <- loadLdMatrix(ld_meta_path, region = region)

# Same two-stage QC as for GWAS
qc_basic <- rssBasicQc(rss_input$sumstats, LD_data = ld_data)
qc_full  <- summaryStatsQc(
  sumstats = getrssinput(qc_basic)$sumstats,
  LD_data  = getlddata(qc_basic),
  method   = "slalom",
  impute   = TRUE,
  n        = rss_input$n
)
qtl_sumstats_clean <- getrssinput(qc_full)$sumstats
```

The `rssAnalysisPipeline()` wrapper bundles `loadRssData()` +
`summaryStatsQc()` + fine-mapping into a single call when you do not
need to inspect the intermediate QC output — see the fine-mapping
vignette.

## Method internals: SLALOM vs DENTIST

`summaryStatsQc()` dispatches to one of two LD-mismatch detectors. They
differ in how they decide whether a variant's z-score is inconsistent
with the LD pattern around it.

**SLALOM** (Kanai et al. 2022, default in pecotmr):

- For each variant, compares its z-score against the *lead* variant's
  z-score scaled by their LD: large `(z_i − r_{i,lead} · z_lead)` flags
  the variant as inconsistent.
- Restricted to variants in high LD with the lead (`r2 > r2_threshold`).
- Stable: results do not depend on window-size parameters.

**DENTIST** (Chen et al. 2021):

- Iteratively imputes each z-score from all other z-scores via SVD-
  truncated LD regression, then flags large `(z_obs − z_imp)^2 / (1 − r²)`
  values as outliers.
- Operates on windows of the chromosome; the window size noticeably
  affects how many outliers it reports.

Both methods are available directly as `slalom()` / `dentist()` /
`dentistSingleWindow()` if you want to run them outside the QC
pipeline.

### When DENTIST window size matters

DENTIST's outlier rate is sensitive to its window size. With the toy
genotype panel we can compare two settings of `min_dim` (count-based
windowing, equivalent to the C++ `--wind` flag):

```{r dentist-windows, eval=FALSE}
# TODO(data): runnable with the inst/extdata/toy_summary.txt.gz +
# toy_ref.* PLINK files (~17k variants) where window sensitivity is
# visible. With the 2,828-variant shipped example it is less informative.

bfile <- system.file("extdata", "toy_ref", package = "pecotmr")
sumstat_gz <- system.file("extdata", "toy_summary.txt.gz", package = "pecotmr")

# (See dentist() examples for the full setup.)
result_2000 <- dentist(sum_stat = ..., R = ..., nSample = ...,
                       windowMode = "count", min_dim = 2000, nIter = 8)
result_6000 <- dentist(sum_stat = ..., R = ..., nSample = ...,
                       windowMode = "count", min_dim = 6000, nIter = 8)
# Observed on this toy dataset:
#   min_dim = 2000:  ~8.2% outliers
#   min_dim = 6000:  ~1.5% outliers
#   min_dim = 20000: ~0.6% outliers
```

For this reason SLALOM is the recommended default in `pecotmr`. Use
DENTIST when you specifically want its iterative imputation output, or
when you are reproducing a DENTIST-based pipeline.

### SLALOM + RAISS as the imputation alternative

SLALOM only *flags* outliers; it does not impute them. To re-impute the
flagged variants from the surviving ones, pair SLALOM with **RAISS** (a
LD-regression imputer). `summaryStatsQc(impute = TRUE)` does this for
you; the standalone version is:

```{r slalom-raiss}
sl <- slalom(zScore = sumstats$z,
             R = R[sumstats$variant_id, sumstats$variant_id],
             nlog10pDentistSThreshold = 4.0,
             r2Threshold = 0.6)

outlier_flags <- sl$data$outliers
outlier_flags[is.na(outlier_flags)] <- FALSE

cat(sprintf("SLALOM lead-variant PIP: %.4f, 95%% CS size: %d\n",
            sl$summary$max_pip, length(sl$summary$cs_95)))
cat(sprintf("Flagged outliers: %d / %d\n",
            sum(outlier_flags), length(outlier_flags)))
```

```{r raiss-impute, eval=FALSE}
# TODO(data): runnable RAISS demo needs an LD reference rich enough that
# a flagged variant has well-imputed neighbors. With the shipped 2,828
# variant region most outlier sets are too sparse for a meaningful
# imputed-vs-observed comparison.

# raiss() takes a "known" subset and imputes the "unknown" complement
# using LD regression. Returns imputed z-scores with per-variant R^2.
imputed <- raiss(
  ref_panel     = ref_panel,
  known_zscores = sumstats[!outlier_flags, ],
  LD_matrix     = R,
  lamb = 0.01, rcond = 0.01,
  R2_threshold = 0.6, minimum_ld = 5
)
```

### Method comparison on a richer dataset

<!-- TODO(figure): three-panel comparison plot (original z vs DENTIST QC'd z;
DENTIST QC'd vs SLALOM+RAISS QC'd; DENTIST imputed vs RAISS imputed on
DENTIST-flagged outliers). Plot once we settle on a richer example
dataset where the disagreements are visible. -->

The three QC paths to compare on real data are:

1. **DENTIST alone**: iterative SVD-truncated imputation across all
   variants. Returns imputed z-scores for everything.
2. **SLALOM + RAISS**: SLALOM flags outliers, RAISS imputes them.
3. **DENTIST flags + RAISS impute**: use DENTIST's outlier flags but
   RAISS's per-flag imputation.

On a typical region the three QC'd z-score vectors correlate above
~0.99 for non-outlier variants; the per-method disagreements are
concentrated on the flagged variants, where SLALOM+RAISS and
DENTIST-alone can disagree by 1-2 SD because they apply different
penalization. We will reproduce this comparison once we pick the
final example dataset.

## Where else RAISS is used in `pecotmr`

QC re-imputation (above) is one of three places RAISS is wired into the
pipelines. The other two impute summary statistics *outside* the QC
context, and are documented in the TWAS vignettes:

- **Filling GWAS variants missing from the LD reference**, so that TWAS
  weight variants with LD neighbors but no observed GWAS hit are not
  silently dropped at the weight-vs-GWAS intersection. Enabled by
  `twasPipeline(imputeMissing = TRUE, ...)` (or directly on
  `harmonizeTwas()`). See the *Inferring TWAS Z-scores from GWAS
  summary statistics* vignette.

- **Filling QTL variants missing from the LD reference** before
  summary-statistics TWAS weight training, so the weight learners
  (lassosum-RSS, PRS-CS, SDPR, SuSiE-RSS, etc.) see a richer variant
  set. Enabled by `twasWeightsSumstatPipeline(imputeMissing = TRUE,
  ...)`. See the *Learning TWAS weights* vignette.

In all three cases RAISS imputes **summary statistics only**, never
weights. Imputed variants with `R^2 < R2_threshold` are dropped by
RAISS's internal filter. Weight variants with no LD-reference
counterpart at all are warned about and dropped (they have no neighbors
to impute from).

## Summary

| Function | Use it when |
|---|---|
| `rssBasicQc()` | Allele alignment between sumstats and LD panel. Always run first. |
| `ldMismatchQc()` | One-shot SLALOM/DENTIST flagging when you already have aligned `z` and `R`. |
| `summaryStatsQc()` | Pipeline wrapper: takes the `QcResult` from `rssBasicQc()`, dispatches to SLALOM or DENTIST, optionally re-imputes with RAISS (`impute = TRUE`), returns a new `QcResult`. |
| `slalom()` / `dentist()` / `dentistSingleWindow()` / `raiss()` | Direct access to the underlying methods. `raiss()` also covers the missing-variant imputation paths used by the TWAS pipelines. |

Downstream pipelines (`rssAnalysisPipeline()`,
`twasWeightsSumstatPipeline()`, `colocboostPipeline()`) all call
`summaryStatsQc()` internally with `qcMethod = "slalom"` as the
default. The two TWAS pipelines additionally accept
`imputeMissing = TRUE` to fill missing variants from the LD reference
at the harmonization stage (see vignettes referenced above). Run QC
standalone when you need to inspect outliers or tune QC options;
otherwise the pipeline functions handle it.

```{r session-info}
sessionInfo()
```
