---
title: "Quality Control for GWAS and QTL Summary Statistics"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Control for GWAS and QTL Summary Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

Regression-with-summary-statistics (RSS) methods — SuSiE-RSS, ColocBoost,
DENTIST, lassosum, PRS-CS, SDPR — all require a summary-statistics table
(`Z` or `BETA`/`SE`) and an LD reference panel that agree about which
variant is which. Two things go wrong in practice:

1. **Allele alignment.** Sumstats and LD panel may flip A1/A2, encode
   strand-ambiguous variants inconsistently, or contain indels and
   duplicates that need filtering.
2. **LD-summary-statistic mismatch.** Even when alleles align, individual
   z-scores can be inconsistent with the LD pattern around them because
   of imputation errors, ancestry differences between the GWAS and LD
   panel, or per-variant sample-size differences.

In the current S4 API both stages are unified behind a single pipeline
entry point:

| Stage | Mechanism |
|---|---|
| Variant-content QC (column standardization, indels, strand-ambiguous, MAF/INFO/N filters, p-value & effect-size sanity checks, optional dbSNP / liftover) | Delegated to `MungeSumstats::format_sumstats()` from inside `summaryStatsQc()`. |
| Pecotmr-specific QC (`skipRegion`, optional PIP screen, panel harmonization against the `ldSketch`, optional SLALOM/DENTIST LD-mismatch QC, optional RAISS imputation) | Implemented inside `summaryStatsQc()`. |

`summaryStatsQc()` is the **only** QC entry point. It takes a
`GwasSumStats` or `QtlSumStats` (a DFrame collection that already carries
its LD reference via the `ldSketch` slot), runs the full pipeline above,
and returns a new collection of the same class with cleaned entries and
a populated `qcInfo` slot. Downstream pipelines refuse SumStats
collections where `length(getQcInfo(x)) == 0L`, so QC is mandatory before
fine-mapping, TWAS-weight learning, colocalization, etc.

```{r load-packages}
library(pecotmr)
```

## Reading summary statistics into pecotmr

There is no `readSumstats()` or `loadRssData()` in the new API. The S4
design is constructor-only: build an in-memory `GRanges` (one per
study/context/trait tuple) carrying the variant metadata in its `mcols`,
then pass a list of those `GRanges` to `GwasSumStats()` or
`QtlSumStats()` together with the LD reference (`ldSketch`).

The minimum required mcols on each entry are `SNP`, `A1`, `A2`, `Z`,
`N`; optional columns are `MAF`, `INFO`, `BETA`, `SE`, `P`. The
`ldSketch` is a `GenotypeHandle` covering the same variant universe
(can be built from a VCF, GDS, PLINK1/2 file, or — new in this release
— from an LD-meta TSV row via `GenotypeHandle(ldMeta = ..., region = ...)`).

### Worked example: GWAS sumstats from a TSV

```{r build-gwas-sumstats, eval=FALSE}
# TODO(data): replace paths with your own; this block is illustrative.

# Read your sumstats with MungeSumstats first to harmonize column names
# and genome build (summaryStatsQc() will re-run it with the curated knobs,
# but it's convenient to standardize before building the GRanges).
df <- MungeSumstats::format_sumstats(
  path             = "your_gwas.tsv.gz",
  ref_genome       = "GRCh38",
  return_data      = TRUE,
  return_format    = "data.frame")

gr <- GenomicRanges::GRanges(
  seqnames = paste0("chr", df$CHR),
  ranges   = IRanges::IRanges(start = df$BP, width = 1L))
S4Vectors::mcols(gr) <- S4Vectors::DataFrame(
  SNP = df$SNP, A1 = df$A1, A2 = df$A2,
  Z   = df$BETA / df$SE,
  N   = df$N)

ldSketch <- GenotypeHandle(ldMeta = "ld_meta.tsv",
                           region = "chr1:1-1000000")

gws <- GwasSumStats(
  study    = "MyStudy",
  entry    = list(gr),
  genome   = "GRCh38",
  ldSketch = ldSketch)
```

### Worked example: QTL sumstats (tabix-indexed per-gene tables)

```{r build-qtl-sumstats, eval=FALSE}
# TODO(data): point at your tabix-indexed per-gene QTL tables and column
# map. The pattern below works for tensorQTL / APEX nominal output once
# columns are renamed to SNP / A1 / A2 / Z / N.

readOneGene <- function(tabix_path, gene_id, region) {
  rows <- Rsamtools::scanTabix(Rsamtools::TabixFile(tabix_path),
                               param = GenomicRanges::GRanges(region))[[1L]]
  df <- read.table(text = rows, sep = "\t",
                   col.names = c("chrom","pos","SNP","A1","A2","BETA","SE","N"))
  df <- df[df$gene_id == gene_id, , drop = FALSE]
  gr <- GenomicRanges::GRanges(
    seqnames = paste0("chr", df$chrom),
    ranges   = IRanges::IRanges(start = df$pos, width = 1L))
  S4Vectors::mcols(gr) <- S4Vectors::DataFrame(
    SNP = df$SNP, A1 = df$A1, A2 = df$A2,
    Z   = df$BETA / df$SE,
    N   = df$N)
  gr
}

ldSketch <- GenotypeHandle(ldMeta = "ld_meta.tsv",
                           region = "chr1:1-1000000")

qts <- QtlSumStats(
  study    = "MyQTL",
  context  = "brain",
  trait    = c("ENSG00000001", "ENSG00000002"),
  entry    = list(readOneGene("brain.tsv.gz", "ENSG00000001", "chr1:1-1000000"),
                  readOneGene("brain.tsv.gz", "ENSG00000002", "chr1:1-1000000")),
  genome   = "GRCh38",
  ldSketch = ldSketch)
```

The binary-trait variance formula `varY = n / (n - 1) * phi * (1 - phi)`
(with `phi = nCase / n`) is computed in user code and passed via the
constructor's `varY = c(<value>)` argument.

## Running the QC

Once the SumStats collection is built, every QC step lives in
`summaryStatsQc()`:

```{r run-qc, eval=FALSE}
gws_qcd <- summaryStatsQc(
  gws,
  useDbsnpRefCheck      = FALSE,   # opt-in to MungeSumstats dbSNP checks
  removeIndels          = FALSE,
  removeStrandAmbiguous = TRUE,
  mafCutoff             = 0,       # 0 = off; > 0 requires MAF column
  infoCutoff            = 0,       # 0 = off; > 0 requires INFO column
  nCutoff               = 5,
  skipRegion            = NULL,    # e.g. MHC "chr6:25000000-35000000"
  zMismatchQc           = "slalom", # "none", "slalom", "dentist"
  alleleFlipKriging     = FALSE,    # opt-in pre-SLALOM LD-consistency check
  impute                = FALSE,    # RAISS imputation against ldSketch
  matchMinProp          = 0)
```

`gws_qcd` is a new `GwasSumStats` of the same class, with cleaned
entries and a populated `qcInfo` slot. The per-entry audit record is
accessible via `getQcInfo()`:

```{r inspect-qcinfo, eval=FALSE}
qc <- getQcInfo(gws_qcd)
str(qc$entryAudit[[1]])    # per-entry counts: variantsIn, mungeSumstatsDropped,
                           # keepVariantsDropped, skipRegionDropped, matchedAgainstSketch,
                           # krigingOutliersDropped, ldMismatchOutliersDropped, ...
qc$options                 # the curated knobs you passed (frozen for provenance)
```

## Column-availability contract

A non-zero `mafCutoff` requires every entry to carry a `MAF` mcol;
non-zero `infoCutoff` requires `INFO`; non-zero `nCutoff` requires `N`.
Missing column with a non-zero cutoff is a hard error — there is no
silent fallback. If you don't want a given filter to run, leave its
cutoff at 0 (its default).

## Method internals: SLALOM vs DENTIST

`summaryStatsQc(zMismatchQc = ...)` dispatches to one of two
LD-mismatch detectors. They differ in how they decide whether a
variant's z-score is inconsistent with the LD pattern around it.

**SLALOM** (Kanai et al. 2022, the recommended default once you turn
zMismatchQc on):

- For each variant, compares its z-score against the *lead* variant's
  z-score scaled by their LD: a large `(z_i − r_{i,lead} · z_lead)`
  flags the variant as inconsistent.
- Restricted to variants in high LD with the lead (`r2Threshold`).
- Stable: results do not depend on window-size parameters.

**DENTIST** (Chen et al. 2021):

- Iteratively imputes each z-score from all other z-scores via SVD-
  truncated LD regression, then flags large
  `(z_obs − z_imp)^2 / (1 − r²)` values as outliers.
- Operates on windows of the chromosome; the window size noticeably
  affects how many outliers it reports.

Both methods are available directly as `slalom()` / `dentist()` /
`dentistSingleWindow()` if you want to run them outside the
`summaryStatsQc()` pipeline (typically to tune parameters or to inspect
the per-variant intermediate output).

### When DENTIST window size matters

DENTIST's outlier rate is sensitive to its window size. On a typical
∼17k-variant test region we observe:

| `min_dim` | Approx outlier rate |
|---|---|
| 2,000 | ∼8 % |
| 6,000 | ∼1.5 % |
| 20,000 | ∼0.6 % |

For this reason SLALOM is the recommended default when you turn
zMismatchQc on. Use DENTIST when you specifically want its iterative
imputation output, or when you are reproducing a DENTIST-based
pipeline.

### SLALOM + RAISS as the imputation alternative

SLALOM only *flags* outliers; it does not impute them. To re-impute the
flagged variants from the surviving ones, pair SLALOM with **RAISS** (a
LD-regression imputer). `summaryStatsQc(zMismatchQc = "slalom", impute = TRUE)`
does this for you. The standalone version is exposed as `raiss()`; the
full set of RAISS knobs lives under the `imputeOpts` argument to
`summaryStatsQc()`:

```{r raiss-args, eval=FALSE}
gws_qcd <- summaryStatsQc(
  gws,
  zMismatchQc = "slalom",
  impute      = TRUE,
  imputeOpts  = list(rcond = 0.01, r2Threshold = 0.6,
                     minimumLd = 5, lamb = 0.01))
```

## Where else RAISS is used

QC re-imputation (above) is one of three places RAISS is wired into the
pipelines. The other two impute summary statistics *outside* the QC
context, and are documented in the TWAS vignettes:

- **Filling GWAS variants missing from the LD reference**, so that TWAS
  weight variants with LD neighbors but no observed GWAS hit are not
  silently dropped at the weight-vs-GWAS intersection. See the
  *Inferring TWAS Z-scores from GWAS summary statistics* vignette.

- **Filling QTL variants missing from the LD reference** before
  summary-statistics TWAS weight training, so the weight learners
  (lassosum-RSS, PRS-CS, SDPR, SuSiE-RSS, etc.) see a richer variant
  set. See the *Learning TWAS weights* vignette.

In all three cases RAISS imputes **summary statistics only**, never
weights. Imputed variants with `R^2 < r2Threshold` are dropped by
RAISS's internal filter. Weight variants with no LD-reference
counterpart at all are warned about and dropped (they have no neighbors
to impute from).

## Summary

| Function | Use it when |
|---|---|
| `summaryStatsQc()` | The single QC entry point. Takes a `GwasSumStats` / `QtlSumStats`, runs MungeSumstats variant-content QC + pecotmr-specific QC (allele harmonization, SLALOM/DENTIST, RAISS), returns a new SumStats with `qcInfo` populated. Mandatory before fine-mapping / TWAS / colocalization. |
| `getQcInfo()` | Inspect the per-entry audit record left behind by `summaryStatsQc()`. |
| `slalom()` / `dentist()` / `dentistSingleWindow()` / `raiss()` | Direct access to the underlying methods for inspection or parameter tuning. |
| `ldMismatchQc()` | One-shot SLALOM/DENTIST flagging when you already have aligned `z` and `R` and want just the per-variant outlier flag. |

Downstream pipelines (`fineMappingPipeline()`, `twasWeightsPipeline()`,
`colocboostPipeline()`, etc.) enforce that `length(getQcInfo(x)) > 0L`
before they accept a SumStats input, so QC is never silently skipped.
The two TWAS pipelines additionally accept `imputeMissing = TRUE` to
fill missing variants from the LD reference at the harmonization stage
(see the TWAS vignettes).

```{r session-info}
sessionInfo()
```
