---
title: "SuSiE wrapper unified top_loci output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SuSiE wrapper unified top_loci output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Purpose

This vignette is the schema-of-record for the unified `top_loci` table that
`pecotmr`'s SuSiE wrapper exposes after the **build_top_loci migration**. It
documents the 22-column shape, the credible-set string convention, the
`(variant, gene, method, cs_membership)` row uniqueness rule, and shows a
worked `postprocess_finemapping_fits()` example that row-binds `susie` and
`susie_inf` rows into a single output.

It is reference memory for the package; downstream callers (xqtl-protocol,
transMap, etc.) consume the same single table.

# Why one table

Before this change the wrapper exposed three artifacts side-by-side:

- `top_loci_long` — one row per (variant, method, coverage, CS membership);
- `top_loci` — a wide table with `pip_<method>` and `CS_<cov>_<method>`
  columns;
- `top_loci_export` — a compact "export" projection built by a separate
  helper.

Three artifacts meant three formatter code paths, three sets of column
naming conventions, and silent inconsistencies between long and wide were
easy to introduce. The user request for this migration was simple:

> "we will not have any build_top_loci_long and build_top_loci_width just a
>  function named build_top_loci. ... there is only one build_top_loci
>  function, no more various formats."

The migration accordingly removes the long/wide/export trio (and their
helpers) and replaces them with a single function
[`build_top_loci()`](../R/susie_wrapper.R) that returns one table. The
top-level `postprocess_finemapping_fits()` calls `build_top_loci()` once
per method per fit and `rbind`s the per-method contributions into the
single final `top_loci` table that `format_finemapping_output()` exposes.

There is no separately exposed long table and no separately exposed wide
table.

# The 22-column schema

The unified `top_loci` table has exactly these columns, in this order:

| #  | Column                  | Type      | Source                                                        |
|---:|-------------------------|-----------|---------------------------------------------------------------|
|  1 | `#chr`                  | character | parsed from `variant`                                          |
|  2 | `start`                 | integer   | `pos - 1` from parsed `variant` (BED 0-based)                   |
|  3 | `end`                   | integer   | `pos` from parsed `variant` (BED half-open)                     |
|  4 | `a1`                    | character | from parsed `variant`                                           |
|  5 | `a2`                    | character | from parsed `variant`                                           |
|  6 | `variant`               | character | the row's variant id in `chr:pos:A2:A1` form                    |
|  7 | `gene`                  | character | `colnames(data_y)[1]` (phenotype identifier)                    |
|  8 | `event`                 | character | `paste(other_quantities$condition_id, gene, sep="_")`          |
|  9 | `n`                     | integer   | `nrow(data_y)` (analyzed sample count)                          |
| 10 | `maf`                   | numeric   | from the `maf` argument                                          |
| 11 | `beta`                  | numeric   | from `sumstats$betahat`                                          |
| 12 | `se`                    | numeric   | from `sumstats$sebetahat`                                        |
| 13 | `pip`                   | numeric   | per-variant SuSiE PIP                                            |
| 14 | `posterior_effect_mean` | numeric   | `colSums(alpha * mu)`                                            |
| 15 | `posterior_effect_se`   | numeric   | `sqrt(pmax(colSums(alpha * mu2) - posterior_effect_mean^2, 0))` |
| 16 | `cs_95`                 | character | per-row CS membership at coverage 0.95 (see below)              |
| 17 | `cs_70`                 | character | per-row CS membership at coverage 0.70                           |
| 18 | `cs_50`                 | character | per-row CS membership at coverage 0.50                           |
| 19 | `cs_95_purity`          | numeric   | 0.95-coverage CS purity for this row (see below)                 |
| 20 | `method`                | character | `"susie"`, `"susie_inf"`, ...                                    |
| 21 | `grange_start`          | integer   | parsed from `other_quantities$region`                            |
| 22 | `grange_end`            | integer   | parsed from `other_quantities$region`                            |

There are no `pip_<method>` or `CS_<cov>_<method>` wide-format columns, no
`top_loci_long` column set (`variant_id` / `coverage` / `cs`), and no
legacy `cs_coverage_0.95` / `PIP` / `variant_ID` / `gene_ID` /
`event_ID` / `region_start` / `region_end` names. Downstream callers
should read the columns above.

## CS membership strings (`cs_95`, `cs_70`, `cs_50`)

The credible-set columns are **character strings**, not integers. Each
value has the form

```
<method>_<cs_index>
```

where `<method>` is the row's `method` column and `<cs_index>` is the
**per-method 1-based** credible-set number. Two consequences:

1. Each method numbers its credible sets independently starting at `1`.
   `susie` may have `susie_1`, `susie_2`, ...; `susie_inf` may have
   `susie_inf_1`, `susie_inf_2`, .... The CS index space is *not*
   sequenced across methods.

2. A variant that is retained by the PIP cutoff but is **not** in any
   credible set at the given coverage carries the sentinel `"<method>_0"`.
   For example, a `susie`-method row with no 0.70-coverage CS membership
   carries `cs_70 = "susie_0"`.

## `cs_95_purity`

A single purity column. It holds the purity of the row's 0.95-coverage
credible set, taken from `cs_tables[[i]]$sets$purity$min.abs.corr` (with a
`cs_corr`-based fallback when `sets$purity` is unavailable). Rows whose
`cs_95` is `"<method>_0"` (i.e. not in any 0.95-coverage CS) carry
`cs_95_purity = 0`. There are no `cs_70_purity` or `cs_50_purity` columns
at this stage.

# Row uniqueness

The row identity is `(variant, gene, method, cs_membership)`. Concretely:

- A variant in a single CS within a single method produces **one** row.
- A variant in CS 1 *and* CS 2 of the same method (overlapping CS) produces
  **two** rows. Both share `(variant, gene, method)` and differ in
  `cs_95` (e.g. one row `cs_95 = "susie_1"`, the other
  `cs_95 = "susie_2"`).
- A variant that appears in CS from more than one method (e.g. `susie` CS 1
  and `susie_inf` CS 1) produces **two** rows. They share `(variant, gene)`
  and differ in `method` (and in `cs_95` because each method numbers
  independently).
- A PIP-only retained variant (no CS membership at any coverage) produces
  **one** row per method with
  `cs_95 = cs_70 = cs_50 = "<method>_0"` and `cs_95_purity = 0`.

This rule preserves overlapping-CS evidence both *within* a method (CS 1
overlaps CS 2 in `susie`) and *across* methods (the same variant appears
in `susie`'s CS and `susie_inf`'s CS).

# Caller responsibilities

`build_top_loci()` reads per-fit metadata from `data_x`, `data_y`, and a
reserved `other_quantities` list. The caller (typically
`univariate_analysis_pipeline()` and through it the xqtl-protocol
`susie_twas` block and the transMap finemapping pipeline) supplies:

| Field                          | Used for                            |
|--------------------------------|-------------------------------------|
| `data_y`                       | `n` and `gene` (column name)        |
| `other_quantities$region`      | `grange_start`, `grange_end`        |
| `other_quantities$condition_id`| `event` composition with `gene`     |

If `other_quantities$region` is missing, `grange_start` and `grange_end`
remain `NA` in the wrapper's output; transMap's
`.transmap_finemap_top_loci_table()` then fills them from its own
analysis-region metadata. If `other_quantities$condition_id` is missing,
`event` falls back to `NA_character_`.

# Worked example

The example below sketches a typical wrapper run that fits both `susie`
and `susie_inf` and shows that the final `top_loci` has one row per
`(variant, gene, method, cs_membership)`.

```{r setup, eval = FALSE}
library(pecotmr)
library(susieR)

set.seed(42)
n <- 300; p <- 50
X <- matrix(rnorm(n * p), n, p)
colnames(X) <- sprintf("chr1:%d:G:A", seq_len(p))
beta <- rep(0, p); beta[c(10, 30)] <- 1.5
y <- as.numeric(X %*% beta) + rnorm(n, sd = 0.5)
Y <- matrix(y, ncol = 1, dimnames = list(NULL, "ENSG00000179403"))

# fit_susie_inf_then_susie() is a pecotmr helper that returns both fits.
fits <- fit_susie_inf_then_susie(X, y)

# Caller-supplied per-fit metadata flows through other_quantities. The two
# reserved subfields are documented contract for the unified export only;
# other_quantities itself remains a free-form list (e.g. dropped_samples
# stays accepted).
oq <- list(
  region       = "chr1:1-50",
  condition_id = "ctx_eQTL"
)

post <- postprocess_finemapping_fits(
  fits, data_x = X, data_y = Y,
  coverage = 0.95, secondary_coverage = c(0.7, 0.5),
  other_quantities = oq
)

out <- format_finemapping_output(post, primary_method = "susie")
tl  <- out$top_loci

# tl is the 22-column unified table. Verify the column order and the
# per-method CS strings.
stopifnot(identical(names(tl), c(
  "#chr", "start", "end", "a1", "a2",
  "variant", "gene", "event",
  "n", "maf", "beta", "se",
  "pip", "posterior_effect_mean", "posterior_effect_se",
  "cs_95", "cs_70", "cs_50", "cs_95_purity",
  "method", "grange_start", "grange_end"
)))
stopifnot(setequal(unique(tl$method), c("susie", "susie_inf")))
stopifnot(all(grepl("^(susie|susie_inf)_\\d+$",
                    c(tl$cs_95, tl$cs_70, tl$cs_50))))
```

A few things to notice in the worked example:

- `out` is what xqtl-protocol / transMap callers receive. The previously
  exposed `out$top_loci_long` field is gone; `out$top_loci` is the only
  top-loci field.
- `tl$method` is one column with multiple values (`"susie"`,
  `"susie_inf"`); there are no per-method `pip_<method>` columns. To
  inspect a single method's rows, filter on `tl$method`.
- `tl$cs_95` etc. are character columns. If you need an integer CS index
  for a particular method, strip the prefix:
  `as.integer(sub("^.*_", "", tl$cs_95[tl$method == "susie"]))`.
- A row whose `cs_95` is `"susie_0"` was retained by the PIP cutoff but is
  not in any 0.95-coverage CS. Such rows have `cs_95_purity == 0` by
  construction.

# Downstream contract

`format_finemapping_output()` exposes exactly one top-loci field,
`top_loci`, with the 22 columns above. The previously exposed
`top_loci_long` and wide `top_loci` (with `pip_<method>` /
`CS_<cov>_<method>` columns) and the standalone `top_loci_export` are all
removed. Non-top-loci fields (`susie_result_trimmed`, `variant_names`,
`context_names`, `analysis_script`, ...) keep their previous semantics.

The `FineMappingResult` S4 slot still satisfies its legacy validity
check (it expects `variant_id` and `method` columns); `susie_wrapper.R`
projects the new `top_loci` into a backward-compatible shape only for the
S4 slot, so `AllClasses.R`, `AllMethods.R`, and `vcf_writer.R` do not need
to change. End-user consumers should read the 22-column `top_loci`
returned by `format_finemapping_output()`, not the slot.
