---
title: "Learning TWAS Weights from Individual-Level Data and Summary Statistics"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning TWAS Weights from Individual-Level Data and Summary Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

TWAS weights are per-variant coefficients that predict a molecular
phenotype (gene expression, splicing, protein abundance) from genotype.
Once learned, the weights are applied to GWAS summary statistics to
infer per-gene TWAS Z-scores (see the
[TWAS Z-score vignette](twas-zscore.html)).

`pecotmr` supports three weight-learning workflows depending on the
data you have:

| Data you have | Entry point | Methods supported |
|---|---|---|
| Individual-level genotype + phenotype, single context | `univariate_analysis_pipeline()` (sets `twas_weights = TRUE`) or `twas_weights_pipeline()` | SuSiE family, mr.ash, elastic net, LASSO, MCP, SCAD, L0Learn, Bayesian alphabet methods |
| Individual-level data, multiple contexts (e.g. multi-tissue) | `twas_multivariate_weights_pipeline()` | mr.mash, mvSuSiE |
| QTL summary statistics, single context | `twas_weights_sumstat_pipeline()` (replaces OTTERS) | lassosum-RSS, PRS-CS, SDPR, SuSiE-RSS family, P+T, MCP-RSS, SCAD-RSS, L0Learn-RSS, mr.ash-RSS |
| QTL summary statistics, multiple contexts | `twas_multivariate_weights_sumstat_pipeline()` | mr.mash-RSS, mvSuSiE-RSS |

All three return a `TWASWeights` S4 object that
`harmonize_twas()` / `twas_pipeline()` consume directly.

```{r load-packages}
library(pecotmr)
```

## Quick start: individual-level weights

The simplest path for individual-level data is to call
`univariate_analysis_pipeline()` with `twas_weights = TRUE` — fine-mapping
and weight learning happen in one call, sharing the fitted SuSiE models.

```{r individual-quickstart}
data(eqtl_region_example)
X <- eqtl_region_example$X
y <- as.numeric(eqtl_region_example$y_res)
maf <- apply(X, 2, function(g) {
  f <- mean(g, na.rm = TRUE) / 2
  min(f, 1 - f)
})

set.seed(42)
sub <- sort(sample(ncol(X), 300))

fit <- univariate_analysis_pipeline(
  X = X[, sub], Y = y, maf = maf[sub],
  twas_weights = TRUE,
  cv_folds = 3,
  finemapping_extra_opts = list(refine = FALSE),
  region = "chr22:32119000-33000000",
  verbose = 0
)

names(fit$twas_weights_result)
getMethodNames(fit$twas_weights_result$twas_weights)
```

The default preset fits ten methods (SuSiE, SuSiE-inf, mr.ash, elastic
net, LASSO, MCP, SCAD, L0Learn, BayesR, BayesC) plus an ensemble
combination learned via stacked regression. The `TWASWeights` S4 object
holds all per-method weight vectors and (optionally) the underlying
fits.

## Available weight methods

Methods are selected by passing a character vector to `weight_methods`
on `twas_weights_pipeline()` or `univariate_analysis_pipeline()`. Short
names map to functions of the form `<method>_weights(stat, LD, ...)`:

| Family | Methods |
|---|---|
| **SuSiE family** | `susie`, `susie_inf`, `susie_ash` (see the [fine-mapping vignette](fine-mapping.html) for what each models) |
| **Mixture priors** | `mrash` (mr.ash adaptive shrinkage) |
| **Penalized regression** | `enet`, `lasso`, `mcp`, `scad`, `l0learn` |
| **Bayesian alphabet** | `bayes_r`, `bayes_l`, `bayes_a`, `bayes_b`, `bayes_c`, `bayes_n`, `b_lasso` |
| **DPR** | `dpr_vb`, `dpr_gibbs`, `dpr_adaptive_gibbs` |
| **Multi-context** | `mrmash`, `mvsusie` (use `twas_multivariate_weights_pipeline()` instead) |

### Presets

| Preset | Methods |
|---|---|
| `"fast_default"` | susie, susie_inf, mrash, enet, lasso, mcp, scad, l0learn |
| `"default"` | fast_default + bayes_r, bayes_c |

```{r presets-example, eval=FALSE}
# Use a preset
fit <- univariate_analysis_pipeline(
  X = X[, sub], Y = y, maf = maf[sub],
  twas_weights = TRUE,
  weight_methods = "fast_default",   # passed through to twas_weights_pipeline
  cv_folds = 3, verbose = 0
)

# Or a custom subset
fit <- twas_weights_pipeline(
  X = X[, sub], y = y,
  susie_fit = fit$susie_fitted,
  weight_methods = c("susie", "enet", "lasso"),
  cv_folds = 3, verbose = 0
)
```

When `weight_methods = "default"` (or `"fast_default"`), the pipeline
also estimates a spike-and-slab sparsity prior from mr.ash and injects
it into BayesC / BayesB before fitting them — set `estimate_pi = FALSE`
to skip.

## Cross-validation and ensemble weights

`cv_folds > 1` runs k-fold CV per method, returning per-fold predictions
plus the metrics needed to filter imputable genes (correlation,
adjusted R², p-value, RMSE, MAE):

```{r cv-output, eval=FALSE}
cv <- fit$twas_weights_result$twas_cv_result
str(cv$performance, max.level = 2)  # method -> matrix(metrics x context)
cv$sample_partition                  # data.frame(Sample, Fold)
```

When `ensemble = TRUE` (default) and at least two methods pass the
`ensemble_r2_threshold` (default 0.01), an additional `ensemble_weights`
method is added by stacked regression over CV predictions. The default
solver is `quadprog`; alternatives are `nnls`, `lbfgsb`, and `glmnet`
(set `ensemble_solver`). The ensemble row is named `ensemble_weights`
in the returned `TWASWeights` object.

### Sample partitioning

`twas_weights_cv()` partitions samples randomly into `cv_folds`. To
control the partition (e.g. to keep related samples in the same fold),
pass a data.frame with `Sample` and `Fold` columns to
`sample_partition`:

```{r sample-partition, eval=FALSE}
sample_partition <- data.frame(
  Sample = rownames(X),
  Fold   = sample(rep(1:5, length.out = nrow(X)))
)
fit <- univariate_analysis_pipeline(
  X = X, Y = y, maf = maf,
  twas_weights = TRUE,
  cv_folds = 5, sample_partition = sample_partition,
  verbose = 0
)
```

### Imputability selection

Downstream TWAS-Z inference (`twas_pipeline()` and `twas_analysis()`)
treats a gene as *imputable* when at least one method passes
`rsq_cutoff` (default 0.01) with `pval < rsq_pval_cutoff` (default
0.05). For each imputable gene the best-performing method's weights are
used. See the [TWAS Z-score vignette](twas-zscore.html) for the
inference layer.

## Multi-context weights

For multi-tissue/multi-context QTL data, fit the multivariate analysis
first (`multivariate_analysis_pipeline()` returns the mr.mash and
mvSuSiE fits), then pass them through
`twas_multivariate_weights_pipeline()`:

```{r multi-context, eval=FALSE}
# TODO(data): replace with real multi-tissue eQTL X and Y matrices and
# data-driven prior matrices computed via your usual workflow.

X_mt <- "<numeric matrix samples x variants>"
Y_mt <- "<numeric matrix samples x contexts>"
prior_matrices <- "<list of data-driven prior matrices>"

# (Step 1) Fit mr.mash and mvSuSiE
mnm_fit <- multivariate_analysis_pipeline(
  X = X_mt, Y = Y_mt,
  data_driven_prior_matrices = prior_matrices
)

# (Step 2) Extract TWAS weights per condition
mt_weights <- twas_multivariate_weights_pipeline(
  X = X_mt, Y = Y_mt, mnm_fit = mnm_fit,
  cv_folds = 5
)
# Returns a list indexed by context name; each element has its own
# twas_weights and twas_cv_result.
```

`mnm_fit` provides the fitted multivariate models;
`twas_multivariate_weights_pipeline()` extracts the per-context weights
that match the contract `harmonize_twas()` expects.

## Summary-statistics weights

When you have QTL summary statistics instead of individual-level data,
use `twas_weights_sumstat_pipeline()`. It replaces the standalone
OTTERS pipeline with an integrated workflow: RSS QC → (optional)
RAISS missing-variant imputation → multi-method weight training.

### Default methods

Out of the box the pipeline trains five methods on each region:

| Method | Origin | Defaults |
|---|---|---|
| **lassosum-RSS** | Mak et al. 2017 | grid over `s = c(0.2, 0.5, 0.9, 1.0)`, lambda |
| **PRS-CS** | Ge et al. 2019 | `phi = 1e-4`, `n_iter = 1000`, burnin 500, thin 5 |
| **SDPR** | Zhou et al. 2023 | `iter = 1000`, `burn = 200`, thin 1 |
| **SuSiE-RSS** | Wang et al. 2020 | normal IBSS |
| **SuSiE-inf-RSS** | Cui et al. 2024 | `unmappable_effects = "inf"`, refine = FALSE |

The SuSiE-RSS and SuSiE-inf-RSS methods share a single two-stage fit
(SuSiE-inf-RSS initialises SuSiE-RSS), so adding both costs little
beyond fitting either alone. See the
[fine-mapping vignette](fine-mapping.html) for the SuSiE-family
comparison.

P+T (pruning and thresholding) weights are also produced for any
`p_thresholds` you pass (default `c(0.001, 0.05)`); set
`p_thresholds = NULL` to skip P+T.

### All available SS-TWAS methods

In addition to the five defaults above, the following are available
through the `methods` argument:

| Method short name | Function | Family / origin |
|---|---|---|
| `lassosum_rss` | `lassosum_rss_weights()` | L1-penalized regression on summary stats (Mak et al. 2017) |
| `prs_cs` | `prs_cs_weights()` | Continuous shrinkage Bayesian prior (Ge et al. 2019) |
| `sdpr` | `sdpr_weights()` | Dirichlet process mixture (Zhou et al. 2023) |
| `susie_rss` | `susie_rss_weights()` | SuSiE-RSS, sparse single effects |
| `susie_inf_rss` | `susie_inf_rss_weights()` | SuSiE with infinitesimal background (`unmappable_effects = "inf"`) |
| `susie_ash_rss` | `susie_ash_rss_weights()` | SuSiE with adaptive shrinkage mixture (`unmappable_effects = "ash"`) |
| `mr_ash_rss` | `mr_ash_rss_weights()` | mr.ash adaptive shrinkage on summary stats |
| `mcp_rss` | `mcp_rss_weights()` | Minimax concave penalty on summary stats |
| `scad_rss` | `scad_rss_weights()` | Smoothly clipped absolute deviation penalty |
| `l0learn_rss` | `l0learn_rss_weights()` | L0Learn best-subset selection on summary stats |
| **P+T** (built-in) | n/a | Marginal-effect thresholding at user-supplied `p_thresholds` |

All `*_rss_weights` functions follow the same contract
`<method>_weights(stat, LD, ...)`: the pipeline computes `stat = list(b
= z / sqrt(n), cor = z / sqrt(n), z = z, n = rep(n, p))` from the QC'd
sumstats and passes the QC'd `LD_mat` to each method. Any function
following that convention can be dropped into `methods` without code
changes — including user-supplied custom methods, as long as the
function returns a per-variant numeric weight vector.

### Quick start

```{r sumstat-quickstart, eval=FALSE}
# TODO(data): replace with real QTL sumstats and LD reference paths
qtl_sumstat_path <- "<path/to/qtl_nominal.tsv.gz>"
qtl_column_map   <- "<path/to/qtl_column_map.yml>"
ld_meta_path     <- "<path/to/ld_meta.tsv>"
region           <- "<chr:start-end>"
gene_id          <- "<ENSG...>"

rss_input <- load_rss_data(
  sumstat_path = qtl_sumstat_path,
  column_file_path = qtl_column_map,
  region = region, extract_region_name = gene_id, region_name_col = 4L
)
ld_data <- load_LD_matrix(ld_meta_path, region = region)

ss_weights <- twas_weights_sumstat_pipeline(
  sumstats = rss_input$sumstats,
  LD_data  = ld_data,
  n        = rss_input$n,
  qc_method = "slalom",       # SLALOM/DENTIST QC; see rss-qc.html
  impute    = TRUE,           # re-impute QC outliers via RAISS
  impute_missing = FALSE      # see next section for missing-variant imputation
)
str(ss_weights, max.level = 2)
```

### Customising methods

To swap in a different method set, pass a `methods` named list. Method
names must end in `_weights` and refer to a function with the contract
`<method>_weights(stat, LD, ...)`:

```{r sumstat-custom, eval=FALSE}
ss_weights <- twas_weights_sumstat_pipeline(
  sumstats = rss_input$sumstats, LD_data = ld_data, n = rss_input$n,
  methods = list(
    lassosum_rss     = list(),
    susie_rss        = list(),
    mr_ash_rss       = list(           # add mr.ash.rss
      var_y = 1, sigma2_e = 0.5,
      s0 = c(0, 0.001, 0.01, 0.1, 0.5),
      w0 = rep(0.2, 5)
    )
  ),
  p_thresholds = c(0.05)
)
```

### Runnable demo on shipped data

For demonstration we feed the shipped GWAS-style sumstats and the
matching genotype reference into the pipeline (a real workflow would
use QTL sumstats and a separate LD panel).

```{r sumstat-demo, eval=FALSE}
data(gwas_sumstats_example)
data(eqtl_region_example)

# Build an LDData object from the shipped genotype panel (in real
# workflows this comes from load_LD_matrix())
X_ref <- eqtl_region_example$X
R <- compute_LD(X_ref, method = "sample")
ref_panel <- parse_variant_id(colnames(X_ref))
ref_panel$variant_id <- colnames(X_ref)
variants_gr <- GenomicRanges::GRanges(
  seqnames = ref_panel$chrom,
  ranges = IRanges::IRanges(start = ref_panel$pos, width = 1L)
)
S4Vectors::mcols(variants_gr) <- S4Vectors::DataFrame(
  variant_id = ref_panel$variant_id, A1 = ref_panel$A1, A2 = ref_panel$A2
)
block_meta <- data.frame(
  block_id = 1L, chrom = ref_panel$chrom[1],
  block_start = min(ref_panel$pos), block_end = max(ref_panel$pos),
  size = ncol(X_ref), start_idx = 1L, end_idx = ncol(X_ref)
)
ld_data <- LDData(correlation = R, variants = variants_gr,
                  block_metadata = block_meta, n_ref = nrow(X_ref))

ss_weights <- twas_weights_sumstat_pipeline(
  sumstats = gwas_sumstats_example,
  LD_data  = ld_data,
  n        = nrow(X_ref),
  methods  = list(susie_rss = list(max_iter = 50)),
  p_thresholds = c(0.05),
  check_ld_method = NULL,
  verbose  = 0
)
getMethodNames(ss_weights$twas_weights)
length(getVariantIds(ss_weights$twas_weights))
```

## Multi-context summary-statistics weights

When you have per-context QTL sumstats (one set per tissue / cell type)
but no individual-level data, use
`twas_multivariate_weights_sumstat_pipeline()`. It mirrors
`twas_multivariate_weights_pipeline()` (individual-level) but works
from sumstats:

1. Run per-context QC (`summary_stats_qc()` with optional SLALOM /
   DENTIST + RAISS re-imputation).
2. Optionally RAISS-impute per-context z-scores for LD-reference
   variants missing from a context.
3. **Align variants across contexts and LD** by intersecting on
   `variant_id` (this is where the alignment happens — passed sumstats
   need only be per-context QC'd; cross-context alignment is automatic).
4. Build the `variants × conditions` `Bhat`, `Shat`, and `Z` matrices.
5. Estimate data-driven prior matrices from the cross-context `Bhat`
   using `compute_cov_flash()` (with `compute_cov_diag()` as fallback) —
   the **same** prior-construction helpers used by the individual-level
   `mrmash_wrapper()` via the shared `build_mrmash_prior_matrices()`.
6. Fit the requested methods (mr.mash-RSS, mvSuSiE-RSS) on the aligned
   matrices.

### Default methods

The pipeline trains two methods by default:

| Method | Function | Underlying call |
|---|---|---|
| **mr.mash-RSS** | `mrmash_rss_weights()` | `mr.mashr::mr.mash.rss()`, sharing prior construction with `mrmash_wrapper()` |
| **mvSuSiE-RSS** | `mvsusie_rss_weights()` | `mvsusieR::mvsusie_rss()`, with `create_mixture_prior()` default |

Both follow the same `<method>_weights(stat, LD, ...)` contract as the
univariate RSS methods, with `stat$z` a variants × conditions matrix
and `stat$n` per-context.

### Quick start

```{r multi-sumstat-quickstart, eval=FALSE}
# TODO(data): replace with real per-context QTL paths
qtl_paths   <- list(
  brain  = "<path/to/brain_qtl.tsv.gz>",
  blood  = "<path/to/blood_qtl.tsv.gz>",
  muscle = "<path/to/muscle_qtl.tsv.gz>"
)
column_map  <- "<path/to/qtl_column_map.yml>"   # shared schema, one file
ld_meta     <- "<path/to/ld_meta.tsv>"
region      <- "<chr:start-end>"
gene_id     <- "<ENSG...>"
n_per_ctx   <- c(brain = 1200, blood = 900, muscle = 500)

# 1. Load per-context sumstats (each tabix-extracted on the same region)
sumstats_list <- lapply(qtl_paths, function(p) {
  rss <- load_rss_data(
    sumstat_path = p, column_file_path = column_map,
    region = region, extract_region_name = gene_id, region_name_col = 4L
  )
  rss$sumstats
})

# 2. Load a shared LD reference for the region
ld_data <- load_LD_matrix(ld_meta, region = region)

# 3. Fit both multivariate RSS methods
mt_ss <- twas_multivariate_weights_sumstat_pipeline(
  sumstats_list = sumstats_list,
  LD_data       = ld_data,
  n             = n_per_ctx,
  methods       = list(
    mrmash_rss  = list(max_iter = 200, verbose = FALSE),
    mvsusie_rss = list(L = 10)
  ),
  qc_method      = "slalom",
  impute         = TRUE,    # QC re-imputation of LD-mismatch outliers
  impute_missing = TRUE     # RAISS imputation of LD variants absent per context
)

# `mt_ss$twas_weights` is a TWASWeights S4 object whose weight matrices are
# (variants x conditions). `mt_ss$qc_summary` reports per-context survivor
# counts and the cross-context intersection size.
str(mt_ss$qc_summary)
```

### Data-driven prior construction

By default, the pipeline estimates the mr.mash data-driven covariance
component from the aligned `Bhat` matrix via
`compute_cov_flash(Bhat)` — the same FLASH-based covariance estimator
the individual-level `mrmash_wrapper()` uses. If FLASH fails (e.g. with
sparse contexts), it falls back to `compute_cov_diag(Bhat)`. Pass
`data_driven_prior_matrices = list(U = ...)` to supply your own
precomputed components, or `canonical_prior_matrices = FALSE` to
disable the canonical mixture.

### Manual prior construction

If you have already built priors elsewhere (e.g. via a separate FLASH
or PCA pass on a training set of effect-size estimates), pass them
directly:

```{r multi-sumstat-priors, eval=FALSE}
# Precomputed prior covariance matrices from a training run
prior_U <- list(
  flash_factor_1 = matrix(...),   # K x K
  flash_factor_2 = matrix(...),
  identity       = diag(K)
)

mt_ss <- twas_multivariate_weights_sumstat_pipeline(
  sumstats_list = sumstats_list, LD_data = ld_data, n = n_per_ctx,
  data_driven_prior_matrices    = list(U = prior_U),
  canonical_prior_matrices      = TRUE,
  estimate_priors_from_sumstats = FALSE,  # bypass the in-pipeline FLASH
  methods = list(mrmash_rss = list())
)
```

### Where alignment happens

Cross-context alignment is **inside** the pipeline at step 3, *after*
per-context QC. Each input `sumstats_list[[k]]` may have a different
variant set; the pipeline runs `summary_stats_qc()` per context, then
intersects survivors across contexts and the LD reference to define the
common variant axis used downstream. To widen the intersection, enable
`impute_missing = TRUE`: per-context RAISS imputes z-scores for LD
variants missing from that context's sumstats *before* the
intersection.

## Missing-variant imputation for SS-TWAS

When the QTL sumstats panel is narrower than the LD reference — common
for nominal eQTL output that filters on per-cohort imputation quality —
weight learners only see variants present in both. Setting
`impute_missing = TRUE` calls RAISS to impute z-scores for LD-reference
variants absent from the QTL sumstats *before* the weight learners
fit, so they see a fuller variant set:

```{r impute-missing, eval=FALSE}
ss_weights_wide <- twas_weights_sumstat_pipeline(
  sumstats = rss_input$sumstats,
  LD_data  = ld_data,
  n        = rss_input$n,
  methods  = list(susie_rss = list(), lassosum_rss = list()),
  qc_method = "slalom",
  impute = TRUE,                                # QC re-imputation (existing)
  impute_missing = TRUE,                        # missing-variant imputation (new)
  impute_opts = list(rcond = 0.01, R2_threshold = 0.6,
                     minimum_ld = 5, lamb = 0.01)
)
```

`impute_missing` is independent of `impute` (the QC re-imputation
option). RAISS imputes summary statistics only — never the weights
themselves. Imputed variants with `R² < R2_threshold` are dropped by
RAISS's internal filter. See the
[RSS QC vignette](rss-qc.html) for the RAISS internals and the
[TWAS Z-score vignette](twas-zscore.html) for the parallel
`impute_missing` option that fills GWAS variants missing from the LD
sketch at TWAS Z-score inference.

## Summary

| Function | Use it when |
|---|---|
| `univariate_analysis_pipeline(twas_weights = TRUE)` | Individual-level, single context. Fine-mapping + weight learning in one call. |
| `twas_weights_pipeline()` | Lower-level individual-level entry. Same internals but takes pre-fitted SuSiE models. |
| `twas_weights_cv()` | Standalone k-fold CV for any subset of methods. |
| `twas_multivariate_weights_pipeline()` | Multi-context individual-level. Wraps mr.mash and mvSuSiE fits. |
| `twas_weights_sumstat_pipeline()` | Single-context sumstats. Replaces standalone OTTERS. Defaults to lassosum-RSS, PRS-CS, SDPR, SuSiE-RSS, SuSiE-inf-RSS. |
| `twas_multivariate_weights_sumstat_pipeline()` | Multi-context sumstats. Per-context QC + cross-context alignment + mr.mash-RSS / mvSuSiE-RSS. |
| `ensemble_weights()` | Stack two or more methods via stacked regression. Called automatically by `twas_weights_pipeline()` when `ensemble = TRUE`. |
| `build_mrmash_prior_matrices()` | Shared prior-matrix builder used by both `mrmash_wrapper()` (individual-level) and `mrmash_rss_weights()` (sumstats). |

All five return a `TWASWeights` S4 object. The
[TWAS Z-score vignette](twas-zscore.html) shows how to apply these
weights to GWAS summary statistics to compute per-gene TWAS Z-scores.

```{r session-info}
sessionInfo()
```
