---
title: "Inferring TWAS Z-Scores from GWAS Summary Statistics"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inferring TWAS Z-Scores from GWAS Summary Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

This vignette picks up where the
[TWAS weights vignette](twas-weights.html) leaves off: a `TwasWeights`
S4 object with per-gene per-variant weights already exists. The
remaining task is to combine those weights with GWAS summary statistics
to obtain a per-gene TWAS Z-score.

`pecotmr` exposes this inference at three layers:

| Layer | Entry point | When to use |
|---|---|---|
| File-based pipeline | `twasPipeline()` | Production: per-gene weights database + tabix-indexed GWAS sumstats + LD sketch genotypes; outputs per-gene-per-context-per-study TWAS results with imputability gating, method fallback, and optional MR. |
| Per-gene inference | `twasAnalysis()` | In-memory: applies many weight methods to one gene's GWAS sumstats and combines per-method p-values (ACAT / HMP / Fisher / Stouffer / etc.) into an omnibus test. |
| Single-method math | `twasZ()`, `twasJointZ()` | When you already have aligned `weights`, `z`, and an LD sketch. Convenient for custom workflows or method comparisons. |

```{r load-packages}
library(pecotmr)
```

## The TWAS Z-score formula

Per single weight vector `w` and GWAS z-score vector `z`, the TWAS test
statistic is

\[
z_\text{TWAS} \;=\; \frac{w^\top z}{\sqrt{w^\top R\, w}}
\]

where `R` is the LD matrix at the weight variants. Under the null,
`z_TWAS` follows a standard normal.

### Eigenvalue / eigenvector reparameterisation

`pecotmr` evaluates the denominator from the **full SVD of the
standardised LD sketch genotype matrix** rather than materialising `R`.
If `X = U D V^\top` is the (untruncated) SVD of the standardised
sketch, then

\[
R \;=\; \frac{X^\top X}{n_\text{sketch}-1}
   \;=\; V\, \Lambda\, V^\top,
\qquad
\Lambda_i \;=\; \frac{D_i^{2}}{n_\text{sketch}-1},
\]

and

\[
w^\top R\, w
\;=\; \sum_{i=1}^{r} \Lambda_i \,\bigl(V^\top w\bigr)_i^{2},
\qquad r \;=\; \mathrm{rank}(X).
\]

The sum runs over **every** nonzero singular value — no top-K
truncation. The two expressions are algebraically equivalent (and
numerically agree to floating-point precision when both are computed),
so this is a reparameterisation, not an approximation.

We use this form for two practical reasons:

- The SVD is computed **once per gene region** and reused across every
  method × condition × study evaluation. Each subsequent denominator
  costs `O(r p)` for `V^\top w` plus `O(r)` for the weighted sum,
  versus `O(p^2)` to multiply against an explicit `R`.
- When the LD sketch has `n_\text{sketch} \le p`, `R` is rank-deficient
  (rank `r`, with `n_\text{sketch} \ge r`). Working in the `r`-dimensional
  eigenbasis represents exactly the information the sketch carries and
  avoids forming a singular `p \times p` matrix.

In practice this means `twasZ()` and `twasJointZ()` consume
`(V, D, n_\text{sketch})` directly from the per-gene LD sketch — `R`
is never formed at any step of the pipeline.

## Single-method inference: `twasZ()`

Given aligned weights, GWAS z-scores, and the LD sketch components,
`twasZ()` returns the per-gene Z and two-sided p-value:

```{r twas-z-demo, eval=FALSE}
# Inputs from the upstream pipelines (twas-weights.html and earlier):
#   w        : numeric vector, one weight per variant for one gene/method
#   gwas_z   : numeric vector of GWAS z-scores, aligned to the same variants
#   V, D, n  : right singular vectors, singular values, sample size of the
#              LD sketch for the gene region

res <- twasZ(weights = w, z = gwas_z,
              V = V, D = D, n_sketch = n)
res$z       # TWAS z-score
res$pval    # two-sided p-value
```

`V`, `D`, and `n_sketch` come from the LD sketch the pipeline already
loads for each gene (see "The file-based pipeline" below); you do not
need to compute them yourself in the normal workflow.

## Combining multiple methods per gene: `twasAnalysis()`

When the gene's `TwasWeights` carries weights for several methods (the
default for both `twasWeightsPipeline()` and
`twasWeightsSumstatPipeline()`), `twasAnalysis()` applies each
method's weights to the GWAS and optionally combines the per-method
p-values into an omnibus test:

```{r multi-method, eval=FALSE}
# weights_matrix: variants x methods (one column per weight method)
# gwas_sumstats_db: data.frame with variant_id, z, ... for the GWAS
# V, D, n_sketch, ld_variant_ids: LD sketch components for the gene region

res <- twasAnalysis(
  weights_matrix       = weights_matrix,
  gwas_sumstats_db     = gwas_sumstats_db,
  extract_variants_objs = rownames(weights_matrix),
  V = V, D = D, n_sketch = n_sketch, ld_variant_ids = ld_variant_ids,
  combine_method       = "acat",
  combine_if_no_cv     = TRUE
)
```

The output is a list keyed by method, each entry containing `z` and
`pval`. When `combine_if_no_cv = TRUE` and at least two methods produce
valid p-values, an `omnibus` entry is appended using the chosen
combine method:

| `combine_method` | Combination |
|---|---|
| `"acat"` (default) | Cauchy combination |
| `"hmp"` | Harmonic mean p-value |
| `"fisher"` | Fisher's method |
| `"stouffer"` | Stouffer's Z |
| `"invchisq"` | Inverse chi-square |
| `"gbj"` | Generalized Berk-Jones |
| `"aspu"` | Adaptive sum of powered scores |
| `"gates"` | GATES gene-based test |

For summary-statistics TWAS where no cross-validation performance is
available to gate methods, `combine_if_no_cv = TRUE` is the standard
default. For individual-level TWAS where CV performance has selected a
best method per gene (`twasPipeline()` handles this automatically),
the omnibus is generally not needed.

## Multi-condition joint test: `twasJointZ()`

For per-condition weights (the output of
`twasMultivariateWeightsPipeline()` or
`twasMultivariateWeightsSumstatPipeline()`), `twasJointZ()`
computes a per-condition TWAS Z plus a cross-condition joint test
using any of the same p-value combination methods as `twasAnalysis()`:

```{r joint, eval=FALSE}
# weights: variants x conditions matrix
# gwas_z:  GWAS z-scores aligned to the rows of `weights`
# V, D_svd, n_sketch: LD sketch SVD for the gene region

joint <- twasJointZ(
  weights        = weights,
  z              = gwas_z,
  V              = V,
  D_svd          = D_svd,
  n_sketch       = n_sketch,
  combine_method = "acat"   # or "hmp", "fisher", "stouffer", "invchisq",
                            # "gbj", "aspu", "gates"
)

joint$Z                          # per-condition z + p-value
joint$combined$method            # the combine method used
joint$combined$pval              # joint cross-condition p-value
```

All eight methods listed in the `twasAnalysis()` table above are
accepted. The cross-condition correlation matrix induced by the weights
and the LD sketch is computed internally and passed to whichever method
needs it (Fisher, Stouffer, inverse chi-square, GBJ, aSPU, GATES);
ACAT and HMP work on the p-values directly.

## The file-based pipeline: `twasPipeline()`

For production use you generally have:

- A weights database produced by `twasWeightsPipeline()` (or the
  summary-statistics analog), saved as one entry per gene-region.
- Tabix-indexed GWAS sumstats files, one per study.
- An LD metadata TSV pointing at LD sketch genotype files (one
  per-chromosome entry, format-agnostic — the loader resolves the
  underlying genotype representation).

`twasPipeline()` consumes all three and produces a
per-gene-per-context-per-study TWAS results table, plus an optional MR
analysis layer.

### Input formats

**`gwas_meta_file`** — a TSV with these columns:

| Column | Meaning |
|---|---|
| `study_id` | Identifier for the GWAS study |
| `chrom` | Integer chromosome the file covers (one row per chromosome per study) |
| `file_path` | Path to the tabix-indexed sumstats file |
| `column_mapping_file` | Path to the YAML column-name map (see [RSS QC vignette](rss-qc.html)) |

**`ld_meta_file_path`** — TSV pointing at the per-gene LD sketch
genotype source. The pipeline loads an LD sketch on the fly for each
gene's window via `loadLdSketch()`; only the sketch path is used —
the full LD matrix is never materialised.

**`twas_weights_data`** — a named list where each entry is a
`TwasWeights` S4 object (one per gene-region). The list keys are
typically gene IDs (`ENSG...`); each entry stores all per-method weight
matrices for that gene's variants. This is the output you carry over
from the [TWAS weights vignette](twas-weights.html).

### Call shape

```{r twas-pipeline, eval=FALSE}
# TODO(data): replace with real weights database, GWAS meta, LD meta
weights_db    <- list(
  ENSG00000179403 = "<TwasWeights S4 object from twasWeightsPipeline>",
  ENSG00000269699 = "<...>"
)
gwas_meta_file <- "<path/to/gwas_meta.tsv>"
ld_meta_path   <- "<path/to/ld_meta.tsv>"
region_block   <- "<chrom_start_end>"  # e.g. "1_1000000_2000000"

twas_res <- twasPipeline(
  twas_weights_data       = weights_db,
  ld_meta_file_path       = ld_meta_path,
  gwas_meta_file          = gwas_meta_file,
  region_block            = region_block,
  ld_reference_sample_size = 17000,         # e.g. ADSP R4
  rsq_cutoff              = 0.01,            # imputability gate (R^2)
  rsq_pval_cutoff         = 0.05,            # imputability gate (p-value)
  mr_method               = "susie",         # MR using SuSiE weights
  mr_pval_cutoff          = 0.05
)

names(twas_res)         # "twas_result", "twas_data", "mr_result"
head(twas_res$twas_result[, c("molecular_id", "context", "gwas_study",
                              "method", "twasZ", "twas_pval",
                              "is_imputable", "is_selected_method")])
```

The pipeline internally:

1. Loads the per-gene LD sketch for the region via `loadLdSketch()`
   and computes its SVD once; only `V`, `D`, and the sketch sample size
   are carried forward — no `p × p` correlation matrix is ever formed.
2. Calls `harmonizeTwas()` to align the weights and GWAS to the LD
   sketch (allele QC, sign-flips, intersection on `variant_id`).
3. Calls `twasAnalysis()` per gene to apply weights to the harmonised
   GWAS — using the LD sketch SVD path through `twasZ()` and including
   any omnibus combination.
4. Applies imputability gating (`rsq_cutoff`, `rsq_pval_cutoff`) and
   method fallback (if the selected method's z-score is NA / non-finite,
   falls back to the next-best by CV R²).
5. Optionally runs an MR layer using `mr_method`'s weights.

### Harmonization steps

The harmonization layer can also be called directly:

- `harmonizeGwas(gwas_file, query_region, ld_variants, col_to_flip,
  match_min_prop, column_file_path, comment_string)` — tabix-loads
  one GWAS file for a region, standardises columns, runs allele QC
  against `ld_variants`, returns the harmonised data.frame.
- `harmonizeTwas(twas_weights_data, ld_meta_file_path, gwas_meta_file,
  ld_reference_sample_size, ..., impute_missing, impute_opts)` —
  loops over genes and studies, loads each LD sketch, calls
  `harmonizeGwas()` per study, then aligns weights to the harmonised
  GWAS via `matchRefPanel()`.

The intersection at the harmonization stage is **by default** between
LD-sketch variants and GWAS variants and weight variants. Weight
variants that have LD-sketch neighbours but no observed GWAS hit get
dropped silently — unless you turn on the option below.

## RAISS imputation of missing GWAS variants

When the GWAS panel is narrower than the LD sketch — common when the
GWAS was imputed with a smaller reference, or when allele QC trims
strand-ambiguous variants — some weight variants get dropped at the
weight-vs-GWAS intersection in `harmonizeTwas()`. Setting
`imputeMissing = TRUE` calls RAISS to impute GWAS z-scores for
LD-sketch variants that are present in the LD sketch but absent
from the GWAS:

```{r impute-missing, eval=FALSE}
twas_res <- twasPipeline(
  twas_weights_data       = weights_db,
  ld_meta_file_path       = ld_meta_path,
  gwas_meta_file          = gwas_meta_file,
  region_block            = region_block,
  ld_reference_sample_size = 17000,
  impute_missing          = TRUE,                # widen GWAS via RAISS
  impute_opts             = list(rcond = 0.01,
                                 R2_threshold = 0.6,
                                 minimum_ld = 5,
                                 lamb = 0.01)
)
```

RAISS imputes **summary statistics only** — never weights. Imputed
variants with `R² < R2_threshold` are dropped. Weight variants that
have no LD-sketch neighbour at all are warned about and dropped; RAISS
cannot impute without neighbours.

## Output table

`twasPipeline()` returns a list with:

- **`twas_result`** — data.frame, one row per gene-context-study-method
  combination, with columns:

  | Column | Meaning |
  |---|---|
  | `chr` | Chromosome |
  | `molecular_id` | Gene ID |
  | `context` | Condition (tissue, cell type) name |
  | `gwas_study` | GWAS study ID from `gwas_meta_file` |
  | `method` | Weight method (e.g. `susie`, `enet`, `lassosumRss`) |
  | `is_imputable` | Whether the gene passed the imputability gate |
  | `is_selected_method` | TRUE for the gene-context's chosen method |
  | `rsq_cv` | Cross-validated R² (individual-level path only) |
  | `pval_cv` | p-value for `rsq_cv` |
  | `twasZ` | TWAS Z-score for this method |
  | `twas_pval` | TWAS p-value |
  | `type` | Molecular trait type (e.g. `expression`, `splicing`) |
  | `block` | LD region identifier |

- **`twas_data`** — list of harmonised gene-context inputs in the shape
  expected by `cTWAS::ctwas_sumstats()`. Only populated when
  `output_twas_data = TRUE`.

- **`mr_result`** — MR analysis output from the chosen `mr_method` and
  coverage cutoff, when MR is enabled.

When `combine_if_no_cv = TRUE` (used in summary-statistics TWAS where
CV performance is not available), the per-method rows are augmented
with an `omnibus` row containing the ACAT-combined p-value.

## Imputability gating and method fallback

Two gates filter the per-gene-context output:

- **Imputability**: a gene-context is kept only if **at least one
  method** has `rsq_cv ≥ rsq_cutoff` and `pval_cv < rsq_pval_cutoff`.
  This is the individual-level CV-based gate. Summary-statistics TWAS
  bypasses this gate (`is_imputable = TRUE` for all genes; no CV
  performance available).
- **Selected method**: within an imputable gene-context, the method
  with the highest `rsq_cv` becomes `is_selected_method = TRUE`. If
  that method's `twasZ` is NA or non-finite, the pipeline falls back
  to the next-best method with a valid z-score, surfacing the swap as
  a `TWAS method fallback` message.

## Summary

| Function | Use it when |
|---|---|
| `twasZ()` | One method × one gene; you already have aligned `weights`, `z`, and the LD sketch SVD `(V, D, n_sketch)`. |
| `twasJointZ()` | Multi-condition weights for one gene; want per-condition TWAS plus a joint test combined via any of `acat`, `hmp`, `fisher`, `stouffer`, `invchisq`, `gbj`, `aspu`, `gates`. |
| `twasAnalysis()` | Multi-method weights for one gene; want per-method TWAS plus an omnibus combination. |
| `twasPipeline()` | Production: weight database + tabix GWAS + LD sketch metadata → per-gene-per-context-per-study results. |
| `harmonizeTwas()` / `harmonizeGwas()` | Lower-level harmonization layer used by `twasPipeline()`. Pass `imputeMissing = TRUE` to widen GWAS via RAISS. |

```{r session-info}
sessionInfo()
```
